---
title: "Stage 1 — IPW Weights"
subtitle: "Inverse Probability Weighting for the PlanningQueues Policy Model"
author: "PlanningQueues / Model"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    highlight: tango
    fig_width: 10
    fig_height: 5
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

pkgs <- c("tidyverse", "lubridate", "broom", "pROC", "kableExtra", "scales", "patchwork")
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p, repos = "https://cloud.r-project.org")
  library(p, character.only = TRUE)
}
```

---

# Purpose

This document is **Stage 1 of the policy model pipeline**. It fits a logistic
model predicting which rows in `ISO_EIA_Merged_v7.csv` have a computable
time-in-queue (TIQ), then produces stabilised inverse probability weights that
correct for the systematic selection bias identified in the missingness
mechanism analysis (AUC = 0.70+, IPW slope divergence = -19.6%).

Every duration model, survival model, and competing-risks model in Stage 2
must use the weights produced here. Rerun this document whenever v5 is updated.

**v2 changes (2026-03-31):** Two fixes informed by `ipw_weight_validation.Rmd`:

1. `iso == "West"` excluded — positivity failure (>50% of West rows had
   p_obs < 0.05 in v1). West utilities have structurally different reporting
   practices; IPW extrapolates rather than corrects for them. Treat as scope
   limitation in all downstream analyses.
2. `iso * queue_yr` interaction added to propensity model — different ISOs
   underwent queue reforms at different times, so observability trends differ
   by ISO. The v1 model's single `queue_yr` coefficient overcorrected the
   time trend (−87.4% slope divergence vs. the expected −19.6%), because it
   conflated selection bias with the trend itself.

**v3 changes (2026-04-01):** Two structural fixes informed by root-cause
diagnosis of the persistent overcorrection (−90.4% slope divergence in v2):

1. **Propensity model training restricted to resolved projects only**
   (`status_group %in% c("came_online", "withdrew")`). The selection bias
   being corrected is: among resolved projects with a known outcome, which
   ones got successfully matched to an EIA completion date? Still-in-queue
   projects have missing TIQ because they haven't resolved yet — that is
   right-censoring, not selection. Including them in the propensity model
   mixes censoring with selection and produces weights that partially
   "correct" for the passage of time rather than for reporting differences.
2. **`queue_yr` removed from propensity formula** → `has_tiq ~ iso + tech + log_mw`.
   Even after restricting to resolved projects, `queue_yr` is both a predictor
   in the propensity model and the covariate driving the trend being estimated
   in Stage 2. Including it makes the IPW partial out year effects before
   Stage 2 can estimate them — causing the −90% overcorrection seen in v2.
   ISO and technology capture the cross-sectional selection differences that
   IPW needs to correct; time trend estimation belongs in Stage 2, not Stage 1.
   West exclusion from v2 is retained.

**Outputs written to `../../data/` (or `$PLANNING_QUEUES_MODEL_DIR` if set):**

- `ipw_weights.csv` — one row per model-eligible input row; join to Stage 2 data on `row_id`
- `ipw_model.rds` — saved logistic model object for inspection and reuse

---

# 1  Load Data

```{r load}
DATA_PATH   <- Sys.getenv(
  "PLANNING_QUEUES_DATA_V7",
  "../../data/ISO_EIA_Merged_v7.csv"
)
MODEL_DIR   <- Sys.getenv("PLANNING_QUEUES_MODEL_DIR", "../../data")
OUT_WEIGHTS <- file.path(MODEL_DIR, "ipw_weights.csv")
OUT_MODEL   <- file.path(MODEL_DIR, "ipw_model.rds")

df_raw <- read_csv(DATA_PATH, show_col_types = FALSE)
```

```{r load-summary, echo=FALSE}
tibble(
  Metric = c("Total rows", "Total columns"),
  Value  = c(comma(nrow(df_raw)), ncol(df_raw))
) %>%
  kbl(col.names = c("", ""), booktabs = TRUE) %>%
  kable_styling(latex_options = c("hold_position"), full_width = FALSE)
```

---

# 2  Feature Engineering

Feature definitions are kept identical to `missingness_mechanism_analysis.Rmd`
so the weights are computed on exactly the same variable specifications.

```{r engineer}
df <- df_raw %>%
  mutate(
    row_id  = row_number(),

    # Missingness flag — outcome of the stage-1 logistic model
    tiq_raw = suppressWarnings(as.numeric(time_in_queue)),
    has_tiq = as.integer(!is.na(tiq_raw) & tiq_raw > 0),

    # Capacity
    mw_num  = suppressWarnings(as.numeric(mw)),
    log_mw  = log1p(mw_num),

    # Queue entry year
    queue_yr = year(suppressWarnings(as.Date(queue_date))),

    # Technology — consistent with all other pipeline scripts
    tech = case_when(
      str_detect(type1, regex("solar|photovoltaic", ignore_case = TRUE))  ~ "Solar",
      str_detect(type1, regex("offshore",           ignore_case = TRUE))  ~ "Offshore Wind",
      str_detect(type1, regex("wind",               ignore_case = TRUE))  ~ "Wind",
      str_detect(type1, regex("batter|storage",     ignore_case = TRUE))  ~ "Battery",
      str_detect(type1, regex("natural gas|gas",    ignore_case = TRUE))  ~ "Gas",
      str_detect(type1, regex("coal",               ignore_case = TRUE))  ~ "Coal",
      str_detect(type1, regex("nuclear",            ignore_case = TRUE))  ~ "Nuclear",
      str_detect(type1, regex("hydro",              ignore_case = TRUE))  ~ "Hydro",
      TRUE                                                                  ~ "Other"
    ),
    tech = factor(tech),
    iso  = factor(iso)
  )
```

```{r engineer-summary, echo=FALSE}
tibble(
  Group   = c("Has computable TIQ", "Missing TIQ", "Missing rate"),
  Value   = c(
    comma(sum(df$has_tiq)),
    comma(sum(df$has_tiq == 0)),
    paste0(round(100 * mean(df$has_tiq == 0), 1), "%")
  )
) %>%
  kbl(col.names = c("", ""), booktabs = TRUE,
      caption = "TIQ availability in full dataset") %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE)
```

---

# 3  Fit Logistic Model

The model predicts `has_tiq` — whether a row has a computable TIQ — from ISO,
technology, log(MW), and queue entry year. Rows with missing predictors are
excluded explicitly here rather than relying on `glm`'s silent dropping, which
would cause length mismatches in downstream `predict()` calls.

```{r model-filter}
# v3: Training set restricted to RESOLVED projects only.
# The selection being corrected is: among projects with a known outcome
# (came_online or withdrew), which ones were successfully matched to an EIA
# completion date? Still-in-queue projects are right-censored — they have
# missing TIQ because they haven't resolved yet, not because of reporting
# differences. Including them in the propensity model conflates censoring
# with selection bias and causes the IPW to upweight recent unresolved rows,
# partially removing the time trend that Stage 2 is meant to estimate.
#
# iso == "West" remains excluded: positivity failure retained from v2
# (>50% of West rows had p_obs < 0.05). West utilities have structurally
# different reporting practices; treat as scope limitation in Stage 2.
model_df <- df %>%
  filter(
    status_group %in% c("completed", "withdrawn"), # <-- Updated to match v5 data
    !is.na(iso),
    !is.na(tech),
    !is.na(log_mw),
    iso != "West"
  ) %>%
  droplevels()
```

```{r model-filter-check, echo=TRUE}
# ── Diagnostic: verify filter produced a usable training set ─────────────────
cat("status_group values in raw data:\n")
print(sort(table(df$status_group, useNA = "ifany")))

cat("\nRows in model_df after filter:", nrow(model_df), "\n")

cat("\niso levels in model_df:\n")
print(levels(model_df$iso))

cat("\ntech levels in model_df:\n")
print(levels(model_df$tech))

# Safety check — stop early with a clear message rather than a cryptic glm error
stopifnot(
  "model_df is empty — check status_group filter values above" = nrow(model_df) > 0,
  "iso has fewer than 2 levels after filtering" = nlevels(model_df$iso) >= 2,
  "tech has fewer than 2 levels after filtering" = nlevels(model_df$tech) >= 2
)
```

```{r model-fit}
# v3 formula: has_tiq ~ iso + tech + log_mw
# queue_yr is intentionally excluded. Even with the training set restricted
# to resolved projects, including queue_yr in the propensity model would
# partial out year effects in the weights before Stage 2 can estimate them —
# because queue_yr is both a predictor here and the covariate driving the
# trend in Stage 2. ISO and technology capture the cross-sectional selection
# differences that IPW needs to correct (SPP ~5% match rate, MISO ~40%).
# Estimating the time trend is Stage 2's job, not Stage 1's.
logit_mod <- glm(
  has_tiq ~ iso + tech + log_mw,
  data   = model_df,
  family = binomial(link = "logit")
)

logit_null <- glm(has_tiq ~ 1, data = model_df, family = binomial())

pseudo_r2 <- 1 - (logit_mod$deviance / logit_null$deviance)
roc_obj   <- roc(logit_mod$y, logit_mod$fitted.values, quiet = TRUE)
auc_val   <- as.numeric(auc(roc_obj))
```

```{r model-diagnostics, echo=FALSE}
tibble(
  Metric  = c("Rows in logistic model", "McFadden pseudo-R²", "AUC", "MCAR verdict"),
  Value   = c(
    comma(nrow(model_df)),
    round(pseudo_r2, 4),
    round(auc_val, 4),
    ifelse(auc_val < 0.55, "Cannot reject MCAR",
    ifelse(auc_val < 0.70, "MCAR rejected — moderate MAR",
                           "MCAR strongly rejected — IPW required"))
  )
) %>%
  kbl(col.names = c("", ""), booktabs = TRUE,
      caption = "Stage-1 logistic model diagnostics") %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE)
```

## 3.1  ROC Curve

```{r fig-roc, echo=FALSE, fig.cap="ROC curve for the missingness prediction model. AUC well above 0.5 confirms MCAR is rejected.", fig.height=4}
roc_df <- data.frame(fpr = 1 - roc_obj$specificities,
                     tpr = roc_obj$sensitivities)

ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_line(linewidth = 1.1, colour = "#0077b6") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  annotate("text", x = 0.65, y = 0.2,
           label = paste0("AUC = ", round(auc_val, 3)),
           fontface = "bold", size = 5, colour = "#0077b6") +
  coord_equal() +
  labs(title = "ROC — Predicting TIQ Observability",
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal(base_size = 12)
```

## 3.2  Coefficient Table

Odds ratios > 1 mean a row with that characteristic is *more* likely to have a
computable TIQ; < 1 means *less* likely (i.e., systematically underrepresented
in the observed sample).

```{r coef-table, echo=FALSE}
coef_tbl <- tidy(logit_mod, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    OR          = round(estimate, 3),
    `95% CI`    = paste0("[", round(conf.low, 3), ", ", round(conf.high, 3), "]"),
    `p-value`   = round(p.value, 4),
    Significant = ifelse(p.value < 0.05, "yes", "")
  ) %>%
  select(Predictor = term, OR, `95% CI`, `p-value`, Significant) %>%
  arrange(`p-value`)

coef_tbl %>%
  kbl(booktabs = TRUE,
      caption = "Logistic model coefficients (OR). OR > 1 = more likely to be observed.") %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"),
                full_width = FALSE) %>%
  row_spec(which(coef_tbl$Significant == "yes"), bold = TRUE) %>%
  row_spec(0, bold = TRUE)
```

---

# 4  Compute Stabilised IPW Weights

The stabilised weight for row $i$ is:

$$w_i = \frac{\bar{p}_{\text{ISO}(i)}}{p_i}$$

where $\bar{p}_{\text{ISO}(i)}$ is the marginal probability of having a
computable TIQ **within the same ISO** as project $i$, and $p_i$ is the fitted
probability from the logistic model. ISO-stratified marginals are used because
match rates differ dramatically across ISOs (SPP \(\approx\)5\%, MISO
\(\approx\)28\%, PJM \(\approx\)98\%). A single pooled marginal would
upweight SPP/MISO relative to PJM purely due to data coverage differences —
not within-ISO selection bias. Stratifying keeps the weights within each ISO
and prevents this compositional shift from dominating Stage 2 estimates.

Weights are trimmed at the 99th percentile to prevent extreme values from
destabilising downstream regressions.

```{r weights}
# v4: ISO-stratified stabilised weights
# Numerator is the per-ISO marginal P(has_tiq), not the pooled marginal.
iso_marginals <- model_df %>%
  group_by(iso) %>%
  summarise(p_marginal_iso = mean(has_tiq), .groups = "drop")

weights_df <- model_df %>%
  left_join(iso_marginals, by = "iso") %>%
  mutate(
    p_obs    = logit_mod$fitted.values,
    ipw_stab = p_marginal_iso / p_obs,
    ipw_trim = pmin(ipw_stab, quantile(ipw_stab, 0.99))
  )

p99_threshold <- quantile(weights_df$ipw_stab, 0.99)
n_extreme     <- sum(weights_df$ipw_stab > 5)
```

```{r weights-summary, echo=FALSE}
tibble(
  Metric = c(
    "P99 trim threshold",
    "Mean weight (trimmed)",
    "Median weight (trimmed)",
    "Max weight (trimmed)",
    "Rows with raw weight > 5"
  ),
  Value = c(
    round(p99_threshold, 3),
    round(mean(weights_df$ipw_trim), 3),
    round(median(weights_df$ipw_trim), 3),
    round(max(weights_df$ipw_trim), 3),
    paste0(n_extreme, " (", round(100 * n_extreme / nrow(weights_df), 1), "%)")
  )
) %>%
  kbl(col.names = c("", ""), booktabs = TRUE,
      caption = "IPW weight summary (v4: ISO-stratified marginals)") %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE)
```

```{r weights-by-iso, echo=FALSE}
# Show per-ISO marginals so it is clear what the numerator is for each ISO
iso_marginals %>%
  arrange(desc(p_marginal_iso)) %>%
  mutate(p_marginal_iso = scales::percent(p_marginal_iso, accuracy = 0.1)) %>%
  kbl(col.names = c("ISO", "Marginal P(has TIQ)"), booktabs = TRUE,
      caption = "Per-ISO marginal match rates used as IPW numerators (v4)") %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE)
```

## 4.1  Weight Distribution

```{r fig-weights, echo=FALSE, fig.cap="Distribution of trimmed IPW weights. Rows with high weights are underrepresented in the observed sample and will be upweighted in Stage 2 models.", fig.height=4}
p_hist <- ggplot(weights_df, aes(x = ipw_trim)) +
  geom_histogram(bins = 60, fill = "#0077b6", colour = "white", linewidth = 0.2) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40") +
  scale_x_continuous(labels = number_format(accuracy = 0.1)) +
  labs(title = "Distribution of Trimmed IPW Weights",
       subtitle = "Dashed line at 1.0 (no reweighting needed). Right tail = underrepresented rows.",
       x = "ipw_trim", y = "Count") +
  theme_minimal(base_size = 11)

p_iso <- weights_df %>%
  group_by(iso) %>%
  summarise(med_w = median(ipw_trim), .groups = "drop") %>%
  filter(!is.na(iso)) %>%
  ggplot(aes(x = reorder(iso, med_w), y = med_w, fill = med_w)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  scale_fill_gradient(low = "#43AA8B", high = "#e63946", guide = "none") +
  labs(title = "Median Weight by ISO",
       subtitle = "ISO > 1 = underrepresented in observed sample",
       x = NULL, y = "Median ipw_trim") +
  theme_minimal(base_size = 11)

p_hist | p_iso
```

```{r fig-weights-tech, echo=FALSE, fig.cap="Median IPW weight by technology. Technologies above 1.0 are underrepresented in the observed sample relative to the full queue population.", fig.height=3.5}
weights_df %>%
  group_by(tech) %>%
  summarise(med_w = median(ipw_trim), n = n(), .groups = "drop") %>%
  filter(!is.na(tech), n >= 20) %>%
  ggplot(aes(x = reorder(tech, med_w), y = med_w, fill = med_w)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  coord_flip() +
  scale_fill_gradient(low = "#43AA8B", high = "#e63946", guide = "none") +
  labs(title = "Median IPW Weight by Technology",
       x = NULL, y = "Median ipw_trim") +
  theme_minimal(base_size = 11)
```

---

# 5  Save Outputs

```{r save}
# Columns to write — row_id is the join key for all Stage 2 models
output_cols <- c(
  "row_id", "iso", "tech", "mw_num", "log_mw", "queue_yr",
  "status_group", "has_tiq", "tiq_raw",
  "p_obs", "ipw_stab", "ipw_trim"
)

if ("queue_id" %in% names(weights_df)) output_cols <- c("queue_id", output_cols)

weights_out <- weights_df %>% select(any_of(output_cols))

# Create output directory if it doesn't exist yet
dir.create(dirname(OUT_WEIGHTS), recursive = TRUE, showWarnings = FALSE)

write_csv(weights_out, OUT_WEIGHTS)
saveRDS(logit_mod, OUT_MODEL)
```

```{r save-summary, echo=FALSE}
n_no_weight <- nrow(df) - nrow(weights_df)

tibble(
  File    = c("ipw_weights.csv", "ipw_model.rds"),
  Rows    = c(comma(nrow(weights_out)), "—"),
  Note    = c(
    paste0(comma(n_no_weight), " input rows excluded (missing predictor) — these cannot be used in IPW-weighted models"),
    "Saved logistic model object for inspection and reuse"
  )
) %>%
  kbl(booktabs = TRUE, caption = paste0("Files written to ", MODEL_DIR)) %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

---

# 6  How to Use These Weights in Stage 2

Join `ipw_weights.csv` to your Stage 2 data on `row_id` and pass `ipw_trim`
as the `weights` argument. Three usage patterns:

```{r usage, eval=FALSE}
library(tidyverse)
library(survival)
library(cmprsk)

weights <- read_csv(file.path(
  Sys.getenv("PLANNING_QUEUES_MODEL_DIR", "../../data"),
  "ipw_weights.csv"
))

model_data <- your_data %>%
  left_join(weights %>% select(row_id, ipw_trim), by = "row_id") %>%
  filter(!is.na(ipw_trim))

# ── Option A: Weighted OLS (baseline, fast) ───────────────────────────────
lm(tiq_years ~ policy_lever + iso + tech + log_mw + queue_yr,
   data    = model_data,
   weights = ipw_trim)

# ── Option B: Weighted parametric survival (recommended for duration) ─────
# Includes still-in-queue rows as right-censored observations.
# event = 1 if came online, 0 if still in queue (censored)
survreg(Surv(tiq_days, event) ~ policy_lever + iso + tech + log_mw + queue_yr,
        data    = model_data %>% filter(status_group != "withdrawn"),
        weights = ipw_trim,
        dist    = "weibull")

# ── Option C: Competing risks (recommended — separates completion vs
#             withdrawal; core model for MW-online objective) ──────────────
# event_code: 1 = came online, 2 = withdrew, 0 = still in queue (censored)
crr(ftime   = model_data$tiq_days,
    fstatus = model_data$event_code,
    cov1    = model_data[, c("policy_lever", "log_mw", "queue_yr",
                              "iso_num", "tech_num")],
    weights = model_data$ipw_trim)
```

---

*Generated: `r Sys.Date()` | Data: `ISO_EIA_Merged_v7.csv` | Stage 1 of PlanningQueues policy model pipeline*
