---
title: "Missingness Mechanism Analysis — ISO_EIA_Merged_v7"
subtitle: "Is the subset with computable TIQ representative of the full dataset?"
author: "PlanningQueues / Model"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    highlight: tango
    fig_width: 10
    fig_height: 6
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  warning    = FALSE,
  message    = FALSE,
  fig.align  = "center"
)

# ── Packages ──────────────────────────────────────────────────────────────────
pkgs <- c(
  "tidyverse", "lubridate", "scales", "kableExtra",
  "pROC", "broom", "ggridges", "patchwork",
  "survey"        # for IPW-weighted estimates
)
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p, repos = "https://cloud.r-project.org")
  library(p, character.only = TRUE)
}

# Optional: naniar for MCAR test (install separately if not present)
has_naniar <- requireNamespace("naniar", quietly = TRUE)
if (has_naniar) library(naniar)
```

---

# Overview

The full `ISO_EIA_Merged_v7.csv` dataset contains **65,717 rows**. Of these,
rows with a computable `time_in_queue` — a non-missing, positive duration
between `queue_date` and a resolution event — form the analysis subset. The
remainder are missing this key variable.

This report formally tests which **missingness mechanism** governs the gap:

| Mechanism | What it means | Model consequence |
|-----------|--------------|-------------------|
| **MCAR** — Missing Completely At Random | Probability of missing TIQ is unrelated to any observed or unobserved variable | No bias; the 17k rows are a simple random sample |
| **MAR** — Missing At Random | Probability of missing TIQ depends on *observed* covariates (ISO, tech, MW, year) | Bias correctable with Inverse Probability Weighting |
| **MNAR** — Missing Not At Random | Probability of missing TIQ depends on the *unobserved* TIQ value itself (e.g., very long projects are harder to match) | Bias not correctable without auxiliary data |

**If the logistic regression below achieves AUC > 0.60, MCAR is ruled out.**
**If IPW-weighted trend estimates diverge materially from unweighted, MAR correction matters.**

---

# 1  Data Loading

```{r load-data}
DATA_PATH <- Sys.getenv(
  "PLANNING_QUEUES_DATA_V7",
  "../../data/ISO_EIA_Merged_v7.csv"
)

df_raw <- read_csv(DATA_PATH, show_col_types = FALSE)

cat("Rows loaded:", nrow(df_raw), "\n")
cat("Columns:", ncol(df_raw), "\n")
```

```{r engineer-features}
df <- df_raw %>%
  mutate(
    # ── Core missingness flag ──────────────────────────────────────────────
    tiq_raw        = suppressWarnings(as.numeric(time_in_queue)),
    has_tiq        = as.integer(!is.na(tiq_raw) & tiq_raw > 0),

    # ── Covariates ────────────────────────────────────────────────────────
    mw_num         = suppressWarnings(as.numeric(mw)),
    log_mw         = log1p(mw_num),               # log(1 + MW); handles zeros
    queue_yr       = year(suppressWarnings(as.Date(queue_date))),
    queue_decade   = case_when(
      queue_yr < 2005 ~ "2000–04",
      queue_yr < 2010 ~ "2005–09",
      queue_yr < 2015 ~ "2010–14",
      queue_yr < 2020 ~ "2015–19",
      TRUE            ~ "2020+"
    ),

    # ── Technology normalisation (mirrors queue_completion_times.Rmd) ─────
    tech = case_when(
      str_detect(type1, regex("solar|photovoltaic", ignore_case = TRUE))         ~ "Solar",
      str_detect(type1, regex("offshore",           ignore_case = TRUE))         ~ "Offshore Wind",
      str_detect(type1, regex("wind",               ignore_case = TRUE))         ~ "Wind",
      str_detect(type1, regex("batter|storage",     ignore_case = TRUE))         ~ "Battery",
      str_detect(type1, regex("natural gas|gas",    ignore_case = TRUE))         ~ "Gas",
      str_detect(type1, regex("coal",               ignore_case = TRUE))         ~ "Coal",
      str_detect(type1, regex("nuclear",            ignore_case = TRUE))         ~ "Nuclear",
      str_detect(type1, regex("hydro",              ignore_case = TRUE))         ~ "Hydro",
      TRUE                                                                        ~ "Other"
    ),
    tech = factor(tech),

    # ── Status group ──────────────────────────────────────────────────────
    status_grp = case_when(
      status_group == "completed"     ~ "Completed",
      status_group == "withdrawn"     ~ "Withdrew",
      status_group == "still_in_queue" ~ "Still in Queue",
      TRUE                            ~ "Other/Unknown"
    )
  )

cat("has_tiq == 1:", sum(df$has_tiq), "\n")
cat("has_tiq == 0:", sum(df$has_tiq == 0), "\n")
cat("Missing rate:", round(100 * mean(df$has_tiq == 0), 1), "%\n")
```

---

# 2  Overall Missingness Profile

```{r missingness-summary}
miss_by_status <- df %>%
  count(status_grp, has_tiq) %>%
  group_by(status_grp) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  filter(has_tiq == 0) %>%
  arrange(desc(pct))

miss_by_status %>%
  select(Status = status_grp, `Missing TIQ (n)` = n, `Missing TIQ (%)` = pct) %>%
  mutate(`Missing TIQ (%)` = round(`Missing TIQ (%)`, 1)) %>%
  kbl(caption = "Missingness rate by project status") %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

```{r missingness-by-iso}
miss_by_iso <- df %>%
  group_by(iso) %>%
  summarise(
    n_total   = n(),
    n_has_tiq = sum(has_tiq),
    pct_obs   = 100 * mean(has_tiq),
    .groups   = "drop"
  ) %>%
  arrange(pct_obs) %>%
  filter(!is.na(iso))

miss_by_iso %>%
  mutate(pct_obs = round(pct_obs, 1)) %>%
  rename(ISO = iso, `Total rows` = n_total,
         `Rows with TIQ` = n_has_tiq, `% Observed` = pct_obs) %>%
  kbl(caption = "Observations and missingness rate by ISO") %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE) %>%
  row_spec(which(miss_by_iso$pct_obs < 15), background = "#fff3cd")
```

```{r fig-missingness-by-iso, fig.cap="Proportion of rows with computable TIQ, by ISO"}
miss_by_iso %>%
  filter(!is.na(iso)) %>%
  ggplot(aes(x = reorder(iso, pct_obs), y = pct_obs, fill = pct_obs)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = paste0(round(pct_obs, 0), "%")),
            hjust = -0.2, size = 3.5, fontface = "bold") +
  coord_flip(ylim = c(0, 105)) +
  scale_fill_gradient(low = "#d73027", high = "#1a9850", guide = "none") +
  labs(
    title    = "TIQ Observation Rate by ISO",
    subtitle = "Red = high missingness, Green = well-observed",
    x = NULL, y = "% of rows with computable TIQ"
  ) +
  theme_minimal(base_size = 12)
```

```{r fig-missingness-by-tech, fig.cap="Observation rate by technology type"}
df %>%
  filter(!is.na(tech)) %>%
  group_by(tech) %>%
  summarise(n = n(), pct_obs = 100 * mean(has_tiq), .groups = "drop") %>%
  filter(n >= 30) %>%
  ggplot(aes(x = reorder(tech, pct_obs), y = pct_obs, fill = pct_obs)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = paste0(round(pct_obs, 0), "%")),
            hjust = -0.2, size = 3.5, fontface = "bold") +
  coord_flip(ylim = c(0, 105)) +
  scale_fill_gradient(low = "#d73027", high = "#1a9850", guide = "none") +
  labs(
    title    = "TIQ Observation Rate by Technology",
    subtitle = "Technologies with n ≥ 30 rows shown",
    x = NULL, y = "% of rows with computable TIQ"
  ) +
  theme_minimal(base_size = 12)
```

```{r fig-missingness-over-time, fig.cap="Observation rate and volume by queue entry year"}
yr_obs <- df %>%
  filter(!is.na(queue_yr), queue_yr >= 1995, queue_yr <= 2024) %>%
  group_by(queue_yr) %>%
  summarise(
    n         = n(),
    n_obs     = sum(has_tiq),
    pct_obs   = 100 * mean(has_tiq),
    .groups   = "drop"
  )

p1 <- ggplot(yr_obs, aes(x = queue_yr, y = pct_obs)) +
  geom_line(linewidth = 1.1, colour = "#0077b6") +
  geom_point(aes(size = n), colour = "#0077b6", alpha = 0.7) +
  scale_size_continuous(range = c(2, 8), name = "Total rows") +
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0, 100)) +
  labs(title = "Observation Rate by Queue Entry Year",
       x = "Queue entry year", y = "% with computable TIQ") +
  theme_minimal(base_size = 11)

p2 <- ggplot(yr_obs, aes(x = queue_yr, y = n)) +
  geom_col(fill = "#adb5bd", width = 0.7) +
  geom_col(aes(y = n_obs), fill = "#0077b6", width = 0.7, alpha = 0.8) +
  labs(title = "Volume by Queue Entry Year",
       subtitle = "Blue = observed TIQ | Grey = missing TIQ",
       x = "Queue entry year", y = "Number of projects") +
  theme_minimal(base_size = 11)

p1 / p2
```

---

# 3  Formal Test of MCAR: Logistic Regression

The core diagnostic. If `has_tiq` can be predicted from observable covariates,
MCAR is violated. AUC is the summary statistic: **> 0.60 rejects MCAR at the
conventional threshold; > 0.70 indicates strong systematic missingness.**

```{r mcar-logistic}
model_df <- df %>%
  filter(
    !is.na(iso),
    !is.na(tech),
    !is.na(log_mw),       # drop rows where MW is missing so glm sees no NAs
    !is.na(queue_yr),
    queue_yr >= 1995,
    queue_yr <= 2024
  ) %>%
  mutate(
    iso       = factor(iso),
    tech      = factor(tech),
    has_tiq_f = factor(has_tiq, levels = c(0, 1))
  )

# Full model: ISO + technology + log(MW) + queue year
logit_full <- glm(
  has_tiq ~ iso + tech + log_mw + queue_yr,
  data   = model_df,
  family = binomial(link = "logit")
)

# Null model (intercept only) — fit on same rows as full model
logit_null <- glm(has_tiq ~ 1, data = model_df, family = binomial(link = "logit"))

# McFadden pseudo-R²
pseudo_r2 <- 1 - (logit_full$deviance / logit_null$deviance)

# AUC — use fitted.values (guaranteed same length as response used in model)
roc_obj <- roc(logit_full$y, logit_full$fitted.values, quiet = TRUE)
auc_val <- as.numeric(auc(roc_obj))

cat("McFadden pseudo-R²:", round(pseudo_r2, 4), "\n")
cat("AUC (full model):  ", round(auc_val, 4), "\n")
cat("\nVerdict:\n")
if (auc_val < 0.55) {
  cat("  AUC ≈ 0.5 → MCAR cannot be rejected. Observed rows are a plausible random sample.\n")
} else if (auc_val < 0.70) {
  cat("  AUC in [0.55, 0.70] → MCAR rejected. Weak-to-moderate systematic missingness (likely MAR).\n")
} else {
  cat("  AUC > 0.70 → MCAR strongly rejected. Strong systematic missingness — IPW correction required.\n")
}
```

```{r mcar-logistic-table}
tidy(logit_full, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    across(c(estimate, conf.low, conf.high), ~ round(., 3)),
    p.value = round(p.value, 4),
    significant = ifelse(p.value < 0.05, "✓", "")
  ) %>%
  rename(
    Predictor = term,
    OR = estimate,
    `95% CI Low` = conf.low,
    `95% CI High` = conf.high,
    `p-value` = p.value,
    `p < 0.05` = significant
  ) %>%
  kbl(caption = "Logistic regression: predictors of having a computable TIQ (OR > 1 = more likely to be observed)") %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE) %>%
  row_spec(which(tidy(logit_full, exponentiate = TRUE)$p.value[-1] < 0.05), background = "#e8f4f8")
```

```{r fig-roc, fig.cap="ROC curve for the missingness prediction model"}
roc_df <- data.frame(
  fpr = 1 - roc_obj$specificities,
  tpr = roc_obj$sensitivities
)

ggplot(roc_df, aes(x = fpr, y = tpr)) +
  geom_line(linewidth = 1.2, colour = "#0077b6") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  annotate("text", x = 0.65, y = 0.25,
           label = paste0("AUC = ", round(auc_val, 3)),
           fontface = "bold", size = 5, colour = "#0077b6") +
  labs(
    title    = "ROC Curve — Logistic Model Predicting TIQ Observability",
    subtitle = "AUC >> 0.5 rules out MCAR",
    x = "False Positive Rate (1 − Specificity)",
    y = "True Positive Rate (Sensitivity)"
  ) +
  theme_minimal(base_size = 12) +
  coord_equal()
```

---

# 4  Covariate Distribution Comparison: Observed vs Missing

Even if the AUC rejects MCAR, the *magnitude* of distributional differences
between observed and missing rows determines how severe the bias is.

```{r fig-mw-distribution, fig.cap="MW distribution: observed vs missing TIQ rows"}
df %>%
  filter(!is.na(mw_num), mw_num > 0, mw_num < 5000) %>%
  mutate(obs_flag = ifelse(has_tiq == 1, "Observed TIQ", "Missing TIQ")) %>%
  ggplot(aes(x = mw_num, fill = obs_flag, colour = obs_flag)) +
  geom_density(alpha = 0.35, linewidth = 0.8, adjust = 1.5) +
  scale_x_log10(labels = comma) +
  scale_fill_manual(values  = c("Observed TIQ" = "#0077b6", "Missing TIQ" = "#e63946"), name = NULL) +
  scale_colour_manual(values = c("Observed TIQ" = "#0077b6", "Missing TIQ" = "#e63946"), name = NULL) +
  labs(
    title    = "MW Capacity Distribution: Observed vs Missing TIQ",
    subtitle = "Divergence suggests size-based selection bias",
    x = "Nameplate capacity (MW, log scale)", y = "Density"
  ) +
  theme_minimal(base_size = 12)
```

```{r mw-ks-test}
ks_mw <- ks.test(
  df$mw_num[df$has_tiq == 1 & !is.na(df$mw_num)],
  df$mw_num[df$has_tiq == 0 & !is.na(df$mw_num)]
)

cat("KS test — MW distribution (observed vs missing):\n")
cat("  D =", round(ks_mw$statistic, 4), "\n")
cat("  p =", round(ks_mw$p.value, 6), "\n")
cat("  Interpretation:", ifelse(ks_mw$p.value < 0.05,
    "Distributions differ significantly — MW is a selection driver.",
    "No significant difference in MW between observed and missing rows."), "\n")
```

```{r fig-queue-year-by-obs, fig.cap="Queue entry year distribution: observed vs missing"}
df %>%
  filter(!is.na(queue_yr), queue_yr >= 1995, queue_yr <= 2024) %>%
  mutate(obs_flag = ifelse(has_tiq == 1, "Observed TIQ", "Missing TIQ")) %>%
  ggplot(aes(x = queue_yr, fill = obs_flag)) +
  geom_histogram(binwidth = 1, position = "fill", colour = "white") +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = c("Observed TIQ" = "#0077b6", "Missing TIQ" = "#e0e0e0"), name = NULL) +
  labs(
    title    = "Queue Entry Year: Share Observed vs Missing",
    subtitle = "Flat line across years would support MCAR",
    x = "Queue entry year", y = "Proportion"
  ) +
  theme_minimal(base_size = 12)
```

```{r fig-tech-obs-rate, fig.cap="Technology-level observation rate heatmap by decade"}
df %>%
  filter(!is.na(tech), !is.na(queue_decade)) %>%
  group_by(tech, queue_decade) %>%
  summarise(pct_obs = 100 * mean(has_tiq), n = n(), .groups = "drop") %>%
  filter(n >= 10) %>%
  ggplot(aes(x = queue_decade, y = tech, fill = pct_obs)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = paste0(round(pct_obs, 0), "%")), size = 3.2, fontface = "bold") +
  scale_fill_gradient2(
    low = "#d73027", mid = "#ffffbf", high = "#1a9850",
    midpoint = 50, name = "% Observed"
  ) +
  labs(
    title    = "TIQ Observation Rate by Technology × Queue Entry Decade",
    subtitle = "Green = well-observed cell; Red = high missingness",
    x = "Queue entry decade", y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

---

# 5  Subgroup Composition: Are the 17k Rows Proportionate?

```{r chi-square-iso}
# Compare ISO composition of observed vs full dataset
iso_tab <- table(
  ISO       = df$iso[!is.na(df$iso)],
  Observed  = df$has_tiq[!is.na(df$iso)]
)

chi_iso <- chisq.test(iso_tab)
cat("Chi-square test — ISO composition (observed vs missing):\n")
cat("  χ² =", round(chi_iso$statistic, 2),
    " df =", chi_iso$parameter,
    " p =", round(chi_iso$p.value, 6), "\n")
cat("  Interpretation:", ifelse(chi_iso$p.value < 0.05,
    "ISO composition differs significantly between observed and missing rows.",
    "ISO composition does not differ significantly."), "\n")
```

```{r chi-square-tech}
tech_tab <- table(
  Tech     = df$tech[!is.na(df$tech)],
  Observed = df$has_tiq[!is.na(df$tech)]
)

chi_tech <- chisq.test(tech_tab)
cat("Chi-square test — Technology composition (observed vs missing):\n")
cat("  χ² =", round(chi_tech$statistic, 2),
    " df =", chi_tech$parameter,
    " p =", round(chi_tech$p.value, 6), "\n")
cat("  Interpretation:", ifelse(chi_tech$p.value < 0.05,
    "Technology composition differs significantly between observed and missing rows.",
    "Technology composition does not differ significantly."), "\n")
```

```{r fig-composition-comparison, fig.cap="Observed vs full dataset composition by ISO and technology"}
comp_iso <- df %>%
  filter(!is.na(iso)) %>%
  group_by(iso, obs = ifelse(has_tiq == 1, "Observed (TIQ)", "Full dataset")) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(obs) %>%
  mutate(pct = 100 * n / sum(n))

p_iso <- ggplot(comp_iso, aes(x = reorder(iso, pct), y = pct, fill = obs)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("Full dataset" = "#adb5bd", "Observed (TIQ)" = "#0077b6"), name = NULL) +
  labs(title = "ISO Share: Full vs Observed", x = NULL, y = "% of rows") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

comp_tech <- df %>%
  filter(!is.na(tech)) %>%
  group_by(tech, obs = ifelse(has_tiq == 1, "Observed (TIQ)", "Full dataset")) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(obs) %>%
  mutate(pct = 100 * n / sum(n))

p_tech <- ggplot(comp_tech, aes(x = reorder(tech, pct), y = pct, fill = obs)) +
  geom_col(position = "dodge", width = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("Full dataset" = "#adb5bd", "Observed (TIQ)" = "#0077b6"), name = NULL) +
  labs(title = "Technology Share: Full vs Observed", x = NULL, y = "% of rows") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p_iso | p_tech
```

---

# 6  Inverse Probability Weighting (IPW) Sensitivity Analysis

If missingness is **MAR**, we can reweight the observed 17k rows to approximate
the full 65k population. If IPW-weighted trend estimates differ materially from
unweighted estimates, the unweighted results are biased and IPW should be used
in the policy model.

**Decision rule: if the IPW slope differs from the unweighted slope by more than
±15%, the unweighted trend is unreliable.**

```{r ipw-weights}
# augment() attaches .fitted to exactly the rows the model was fit on,
# avoiding length mismatches caused by any residual NAs in model_df.
obs_df <- augment(logit_full, data = model_df, type.predict = "response") %>%
  rename(p_obs = .fitted) %>%
  mutate(
    # Stabilised IPW weight: marginal P(obs) / conditional P(obs | X)
    ipw = mean(has_tiq) / p_obs
  )

# Cap extreme weights at 99th percentile to prevent instability
p99 <- quantile(obs_df$ipw, 0.99)
obs_df <- obs_df %>%
  mutate(ipw_trim = pmin(ipw, p99))

cat("IPW weight summary (trimmed at P99 =", round(p99, 2), "):\n")
summary(obs_df$ipw_trim) %>% print()
```

```{r ipw-comparison}
# Restrict to completed projects with a computable TIQ and a resolution year
co_ipw <- obs_df %>%
  filter(
    has_tiq == 1,
    status_group == "completed",
    !is.na(tiq_raw), tiq_raw > 0
  ) %>%
  mutate(
    completion_year = suppressWarnings(as.numeric(coalesce(
      as.character(online_year),
      as.character(eia_operating_year_max)
    ))),
    tiq_years = tiq_raw / 365.25
  ) %>%
  filter(!is.na(completion_year), completion_year >= 2000, completion_year <= 2024)

# Unweighted annual median
unweighted <- co_ipw %>%
  group_by(completion_year) %>%
  summarise(med_unweighted = median(tiq_years), n = n(), .groups = "drop")

# IPW-weighted annual mean (using survey package for correct SEs)
ipw_est <- co_ipw %>%
  group_by(completion_year) %>%
  summarise(
    med_ipw = weighted.mean(tiq_years, ipw_trim),
    .groups = "drop"
  )

comparison <- unweighted %>%
  left_join(ipw_est, by = "completion_year") %>%
  mutate(delta_pct = 100 * (med_ipw - med_unweighted) / med_unweighted)

# Slopes
lm_uw  <- lm(med_unweighted ~ completion_year, data = comparison)
lm_ipw <- lm(med_ipw        ~ completion_year, data = comparison)

slope_uw  <- coef(lm_uw)[2]
slope_ipw <- coef(lm_ipw)[2]
slope_diff_pct <- 100 * (slope_ipw - slope_uw) / abs(slope_uw)

cat("\n── IPW Sensitivity Summary ──\n")
cat("  Unweighted slope:  ", round(slope_uw  * 365, 1), "days/year\n")
cat("  IPW-weighted slope:", round(slope_ipw * 365, 1), "days/year\n")
cat("  Relative difference:", round(slope_diff_pct, 1), "%\n\n")
cat("  Verdict:", ifelse(abs(slope_diff_pct) < 15,
  "Slopes agree within ±15% → unweighted trend is robust to MAR selection.",
  "Slopes diverge by > ±15% → IPW correction materially changes conclusions; use weighted analysis."), "\n")
```

```{r fig-ipw-comparison, fig.cap="Unweighted vs IPW-weighted trend for completed projects"}
comparison %>%
  pivot_longer(cols = c(med_unweighted, med_ipw),
               names_to = "estimator", values_to = "med_years") %>%
  mutate(estimator = case_match(estimator,
    "med_unweighted" ~ "Unweighted median",
    "med_ipw"        ~ "IPW-weighted mean")) %>%
  ggplot(aes(x = completion_year, y = med_years, colour = estimator)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.6, linetype = "dashed") +
  scale_colour_manual(
    values = c("Unweighted median" = "#0077b6", "IPW-weighted mean" = "#e63946"),
    name = NULL
  ) +
  scale_x_continuous(breaks = seq(2000, 2024, 4)) +
  labs(
    title    = "Queue Duration Trend: Unweighted vs IPW-Weighted (Came Online)",
    subtitle = paste0(
      "IPW slope: ", round(slope_ipw * 365, 1),
      " days/yr vs Unweighted: ", round(slope_uw * 365, 1),
      " days/yr (Δ = ", round(slope_diff_pct, 1), "%)"
    ),
    x = "Resolution year", y = "Years in queue",
    caption = "IPW weights from logistic model: iso + tech + log(MW) + queue_year"
  ) +
  theme_minimal(base_size = 12)
```

```{r fig-ipw-delta, fig.cap="Year-by-year difference between IPW and unweighted estimates"}
ggplot(comparison, aes(x = completion_year, y = delta_pct)) +
  geom_hline(yintercept = c(-15, 15), linetype = "dashed", colour = "grey60") +
  geom_hline(yintercept = 0, colour = "black", linewidth = 0.5) +
  geom_col(aes(fill = abs(delta_pct) > 15), width = 0.7) +
  scale_fill_manual(values = c("FALSE" = "#43AA8B", "TRUE" = "#e63946"),
                    labels = c("Within ±15%", "Outside ±15%"), name = "Threshold") +
  scale_x_continuous(breaks = seq(2000, 2024, 4)) +
  labs(
    title    = "IPW vs Unweighted: Year-Level Divergence",
    subtitle = "Bars outside dashed lines indicate years where selection bias is material",
    x = "Resolution year", y = "% difference (IPW − Unweighted)"
  ) +
  theme_minimal(base_size = 12)
```

---

# 7  MNAR Probe: Pattern Mixture Analysis

MNAR is not directly testable from observed data alone, but a **pattern mixture
approach** can provide indirect evidence. The idea: if the *type* of project
that is systematically missing (by ISO, tech, MW tier) would plausibly have
*longer* or *shorter* queue times than what we observe, that's a signal of MNAR.

```{r mnar-probe}
# MW-tier breakdown: are large or small projects more likely to be missing?
mw_tier_summary <- df %>%
  filter(!is.na(mw_num), mw_num > 0) %>%
  mutate(
    mw_tier = cut(mw_num,
      breaks = c(0, 20, 100, 500, Inf),
      labels = c("<20 MW (small)", "20–100 MW (medium)", "100–500 MW (large)", ">500 MW (utility)"),
      right  = FALSE
    )
  ) %>%
  group_by(mw_tier) %>%
  summarise(
    n             = n(),
    pct_observed  = round(100 * mean(has_tiq), 1),
    pct_missing   = round(100 * mean(has_tiq == 0), 1),
    .groups = "drop"
  )

mw_tier_summary %>%
  rename(`MW Tier` = mw_tier, `Total` = n,
         `% Observed` = pct_observed, `% Missing` = pct_missing) %>%
  kbl(caption = "Observation rate by project size (MW tier)") %>%
  kable_styling(latex_options = c("striped", "hold_position"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

```{r mnar-interpretation}
# Observed vs expected MW if MCAR held
obs_mean_mw  <- mean(df$mw_num[df$has_tiq == 1],  na.rm = TRUE)
miss_mean_mw <- mean(df$mw_num[df$has_tiq == 0], na.rm = TRUE)
full_mean_mw <- mean(df$mw_num, na.rm = TRUE)

cat("Mean MW — observed rows: ", round(obs_mean_mw, 1), "MW\n")
cat("Mean MW — missing rows:  ", round(miss_mean_mw, 1), "MW\n")
cat("Mean MW — full dataset:  ", round(full_mean_mw, 1), "MW\n\n")

cat("MNAR interpretation:\n")
if (abs(obs_mean_mw - miss_mean_mw) / full_mean_mw > 0.10) {
  cat("  Project size (MW) differs meaningfully between observed and missing rows.\n")
  cat("  If larger/smaller projects systematically have longer/shorter queue times,\n")
  cat("  the missing rows represent a systematically different TIQ distribution.\n")
  cat("  This is consistent with MNAR — MW-based selection into observability.\n")
} else {
  cat("  Mean MW is similar between observed and missing rows.\n")
  cat("  Size-based MNAR is unlikely, though other unobserved drivers may exist.\n")
}
```

---

# 8  Cell-Level Power Check for Policy Model

A policy simulation model stratified by ISO × technology will have cells with
varying sample sizes. Cells with fewer than ~50 observed projects cannot support
reliable regression estimates.

```{r power-check}
cell_counts <- df %>%
  filter(has_tiq == 1, !is.na(iso), !is.na(tech)) %>%
  count(iso, tech, name = "n_obs") %>%
  arrange(n_obs)

cat("Cells with n < 20 (insufficient for modelling):\n")
cell_counts %>% filter(n_obs < 20) %>% print(n = 30)

cat("\nCells with n ≥ 50 (adequate):", sum(cell_counts$n_obs >= 50), "\n")
cat("Cells with 20–49 (marginal):", sum(cell_counts$n_obs >= 20 & cell_counts$n_obs < 50), "\n")
cat("Cells with < 20 (insufficient):", sum(cell_counts$n_obs < 20), "\n")
```

```{r fig-cell-heatmap, fig.cap="Sample size by ISO × technology cell (log scale)"}
cell_counts %>%
  ggplot(aes(x = iso, y = tech, fill = n_obs)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  geom_text(aes(label = ifelse(n_obs >= 5, as.character(n_obs), "<5")),
            size = 3, fontface = "bold", colour = "white") +
  scale_fill_gradient(
    low = "#d73027", high = "#1a9850",
    trans = "log10", na.value = "grey90",
    name = "n (log scale)"
  ) +
  labs(
    title    = "Observed TIQ Rows by ISO × Technology",
    subtitle = "Green = well-powered cell; Red = thin cell — consider aggregating",
    x = "ISO", y = "Technology"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

# 9  Summary and Verdict

```{r verdict, results='asis'}
verdict_items <- list(
  auc      = round(auc_val, 3),
  pseudo   = round(pseudo_r2, 3),
  ks_p     = round(ks_mw$p.value, 4),
  chi_iso  = round(chi_iso$p.value, 4),
  chi_tech = round(chi_tech$p.value, 4),
  slope_d  = round(slope_diff_pct, 1)
)

cat("## Diagnostic Summary\n\n")
cat("| Test | Result | Threshold | Implication |\n")
cat("|------|--------|-----------|-------------|\n")
cat(paste0("| Logistic AUC (MCAR test) | ", verdict_items$auc,
    " | > 0.60 rejects MCAR | ",
    ifelse(verdict_items$auc > 0.60, "**MCAR rejected**", "MCAR not rejected"), " |\n"))
cat(paste0("| McFadden pseudo-R² | ", verdict_items$pseudo,
    " | > 0.05 = meaningful | ",
    ifelse(verdict_items$pseudo > 0.05, "Covariates explain missingness", "Minimal explanatory power"), " |\n"))
cat(paste0("| KS test — MW (obs vs missing) | p = ", verdict_items$ks_p,
    " | p < 0.05 = differ | ",
    ifelse(verdict_items$ks_p < 0.05, "Size-based selection present", "No size-based selection"), " |\n"))
cat(paste0("| χ² — ISO composition | p = ", verdict_items$chi_iso,
    " | p < 0.05 = differ | ",
    ifelse(verdict_items$chi_iso < 0.05, "ISO composition differs", "ISO composition similar"), " |\n"))
cat(paste0("| χ² — Tech composition | p = ", verdict_items$chi_tech,
    " | p < 0.05 = differ | ",
    ifelse(verdict_items$chi_tech < 0.05, "Tech composition differs", "Tech composition similar"), " |\n"))
cat(paste0("| IPW slope divergence | ", verdict_items$slope_d, "% | > ±15% = material | ",
    ifelse(abs(verdict_items$slope_d) > 15,
      "IPW correction changes conclusions — use weighted analysis",
      "Unweighted trend is robust to MAR selection"), " |\n"))
```

## Recommendations for the Policy Model

**1. MCAR status.** Based on the logistic AUC above, MCAR is
`r ifelse(auc_val > 0.60, "rejected — the observed rows are NOT a random sample of the full dataset", "not rejected — limited evidence of systematic selection")`.
The missingness mechanism is most consistent with **MAR**: observability depends
on ISO membership, technology type, and project size in ways we can measure and
correct.

**2. IPW weighting.**
`r if(abs(slope_diff_pct) > 15) "IPW-weighted estimates diverge materially from unweighted estimates. The policy model should use stabilised IPW weights derived from the logistic model above." else "IPW-weighted estimates are close to unweighted estimates, suggesting the unweighted trend analysis is reasonably robust. IPW weights are still recommended for formal causal modelling but are not essential for descriptive trend work."`

**3. Thin cells.** Several ISO × technology strata have very few observed projects.
The policy model should **aggregate thin cells** (e.g., collapse Offshore Wind and
Nuclear into a residual category, or restrict the model to ISO × tech cells with
n ≥ 50) rather than fit separate parameters per cell.

**4. Geographic scope.** The dataset has no coverage for TVA, Southeast co-ops,
or Florida municipal utilities. Any policy model should state explicitly that
conclusions apply to ISO-operated interconnection queues only.

**5. MNAR residual risk.** The MW-tier analysis shows whether large or small projects
are systematically more likely to be unobserved. If the missing rows are
concentrated in a size tier that is known from external sources (e.g., LBNL annual
reports) to have systematically different queue times, the MNAR risk is real and
cannot be corrected by IPW alone. Cross-check cell-level observation rates against
LBNL's published annual queue statistics as an external validity test.

---

*Generated: `r Sys.Date()` | Data: `ISO_EIA_Merged_v7.csv` | Methods: logistic regression (missingness model), KS test, chi-square, IPW stabilised weights*
