---
title: "Stage 2 — Policy Model"
subtitle: "Competing-Risks Duration Model of US Interconnection Queue Outcomes"
author: "PlanningQueues / Model"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    highlight: tango
    fig_width: 10
    fig_height: 5
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

pkgs <- c("tidyverse", "lubridate", "survival", "cmprsk",
          "broom", "kableExtra", "scales", "patchwork")
for (p in pkgs) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p, repos = "https://cloud.r-project.org")
  library(p, character.only = TRUE)
}
```

---

# Purpose

This document is **Stage 2 of the policy model pipeline**. It estimates the
effect of policy levers on US interconnection queue outcomes using a
competing-risks duration framework. Projects can either come online (event 1)
or withdraw (event 2); the two are competing failures, not censoring of each
other. All models are weighted by `ipw_trim` from Stage 1.

**Three objectives, in order of priority:**

1. **Reduce time to come online** — how much faster can projects complete
   interconnection under different policy regimes?
2. **Reduce withdrawals** — can more projects survive to completion?
3. **Maximise MW coming online** — with a preference for green technologies
   (Solar, Wind, Battery, Offshore Wind).

**Prerequisites:** Run `compute_ipw_weights.Rmd` first. This document reads
`ipw_weights.csv` from `../../data/` (or `$PLANNING_QUEUES_MODEL_DIR`).

---

# Scope Limitations and IPW Notes

These limitations apply to all results in this document and must be stated in
any publication or presentation of findings.

## Data coverage

- **ISO-operated queues only.** TVA, Southeast co-ops, Florida municipals, and
  other non-ISO utilities are excluded. The `iso == "Southeast"` grouping
  aggregates several smaller ISOs and should be interpreted with caution.
- **West excluded.** WAPA, BPA, and other West-of-WECC utilities have
  structurally different reporting practices. The Stage 1 propensity model
  cannot reliably reweight them (positivity failure: >50% of rows had
  p_obs < 0.05). West projects are excluded from all weighted analyses.
- **Entry years before 2008 are unreliable.** ISO queue capture rates are
  below 50% before 2008. Analyses should be anchored to 2008+ cohorts;
  2000–2007 data is retained for trend context only.
- **ERCOT completion dates are sparse.** ERCOT's GIM report is a snapshot only
  — 124 of 576 completed ERCOT projects have an Actual Completion Date.
  ERCOT results are included but should be flagged as underestimates.

## IPW limitations

The Stage 1 IPW weights correct for **within-ISO selection bias by technology
and project size**. They do not — and cannot — correct for the following:

- **Cross-ISO EIA-860 coverage rates.** SPP has a ~5% match rate; MISO ~28%.
  These low rates reflect EIA-860 data coverage, not within-ISO selection.
  There is no basis to assume the unmatched SPP/MISO completed projects have
  different queue time dynamics than the matched 5–28%. IPW cannot reweight
  across this gap without essentially fabricating data.
  **Mitigation:** ISO fixed effects in all Stage 2 models absorb cross-ISO
  composition differences. SPP and MISO results should be treated as lower
  confidence than PJM, NEISO, NYISO, CAISO.
- **Unobserved confounders.** Developer financial resources, local permitting
  difficulty, transmission congestion severity, and interconnection study
  complexity are not in the data. Residual bias from these factors cannot be
  eliminated by IPW and should be acknowledged as a limitation.
- **Weight magnitude.** ISO-stratified IPW weights range from ~0.5 to 2.7
  (max P99-trimmed). The modest range reflects that within-ISO selection by
  tech and MW is real but not severe. The weights shift model estimates
  slightly; they are not driving the results.

```{r ipw-scope-note, echo=FALSE}
message("IPW scope note printed to console at knit time — see Purpose section.")
```

---

# 1  Load Data

```{r paths}
DATA_PATH    <- Sys.getenv(
  "PLANNING_QUEUES_DATA_V7",
  "../../data/ISO_EIA_Merged_v7.csv"
)
MODEL_DIR    <- Sys.getenv("PLANNING_QUEUES_MODEL_DIR", "../../data")
WEIGHTS_PATH <- file.path(MODEL_DIR, "ipw_weights.csv")
OUT_DIR      <- MODEL_DIR
```

```{r load}
df_raw  <- read_csv(DATA_PATH,    show_col_types = FALSE)
weights <- read_csv(WEIGHTS_PATH, show_col_types = FALSE)
```

```{r engineer}
df <- df_raw %>%
  mutate(
    row_id   = row_number(),
    tiq_raw  = suppressWarnings(as.numeric(time_in_queue)),
    has_tiq  = as.integer(!is.na(tiq_raw) & tiq_raw > 0),
    mw_num   = suppressWarnings(as.numeric(mw)),
    log_mw   = log1p(mw_num),
    queue_yr = year(suppressWarnings(as.Date(queue_date))),
    tech = case_when(
      str_detect(type1, regex("solar|photovoltaic", ignore_case = TRUE)) ~ "Solar",
      str_detect(type1, regex("offshore",           ignore_case = TRUE)) ~ "Offshore Wind",
      str_detect(type1, regex("wind",               ignore_case = TRUE)) ~ "Wind",
      str_detect(type1, regex("batter|storage",     ignore_case = TRUE)) ~ "Battery",
      str_detect(type1, regex("natural gas|gas",    ignore_case = TRUE)) ~ "Gas",
      str_detect(type1, regex("coal",               ignore_case = TRUE)) ~ "Coal",
      str_detect(type1, regex("nuclear",            ignore_case = TRUE)) ~ "Nuclear",
      str_detect(type1, regex("hydro",              ignore_case = TRUE)) ~ "Hydro",
      TRUE                                                                 ~ "Other"
    ),
    tech = factor(tech, levels = c("Solar", "Wind", "Battery", "Offshore Wind",
                                   "Gas", "Coal", "Nuclear", "Hydro", "Other")),
    iso  = factor(iso),

    # ── Event codes for competing-risks framework ─────────────────────────────
    # 0 = still in queue (right-censored — no event observed)
    # 1 = came online    (completion event)
    # 2 = withdrew       (competing failure)
    event_code = case_when(
      status_group == "completed"  ~ 1L,
      status_group == "withdrawn"  ~ 2L,
      TRUE                         ~ 0L   # active, eia_only, or unknown
    ),

    # Duration: use tiq_raw if available; for right-censored rows we still
    # need a duration — use days from queue_date to the data snapshot.
    # tiq_raw is in days.
    tiq_days = tiq_raw
  )
```

```{r join-weights}
# Join weights; rows without weights (West, missing predictors) will have
# NA ipw_trim and are excluded from weighted models below.
model_df <- df %>%
  left_join(weights %>% select(row_id, p_obs, ipw_stab, ipw_trim), by = "row_id") %>%
  filter(
    !is.na(ipw_trim),         # must have valid weight
    iso != "West",            # positivity failure — excluded from all analyses
    !is.na(tiq_days),
    tiq_days > 0,
    !is.na(queue_yr),
    !is.na(mw_num),
    !is.na(tech)
  ) %>%
  droplevels()

# Restrict to ≥2008 for primary analyses (capture rate < 50% before then)
model_df_main <- model_df %>% filter(queue_yr >= 2008)
```

```{r load-summary, echo=FALSE}
tibble(
  Metric = c("Total rows in v5", "Rows with valid weights (non-West)",
             "Rows in primary analysis (≥2008)",
             "Event 1 — came online", "Event 2 — withdrew",
             "Event 0 — right-censored"),
  Value = c(
    comma(nrow(df_raw)),
    comma(nrow(model_df)),
    comma(nrow(model_df_main)),
    comma(sum(model_df_main$event_code == 1)),
    comma(sum(model_df_main$event_code == 2)),
    comma(sum(model_df_main$event_code == 0))
  )
) %>%
  kbl(col.names = c("", ""), booktabs = TRUE,
      caption = "Dataset summary for Stage 2") %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE)
```

---

# 2  Policy Levers

Three levers are constructed. Each captures a distinct regulatory mechanism
that theory predicts should affect queue duration and withdrawal rates.

## 2.1  Queue Congestion at Entry

A project entering a congested queue faces longer study timelines because
interconnection studies are processed sequentially (cluster or first-come
first-served). We measure congestion as the number of projects that were
**active in the queue** at the time a given project filed — i.e., projects
that had entered before it and had not yet resolved.

```{r lever-congestion}
# For each project i with queue entry date d_i:
# congestion_i = count of projects j where queue_date_j < queue_date_i
#                and (resolution_date_j > d_i OR resolution_date_j is NA)
#
# Approximation: we use queue_yr as entry year and count same-year or
# earlier-entry projects still in queue at project i's entry year.
# This is a cohort-level approximation; row-level date matching would require
# the raw date field which may have quality issues.

congestion_by_yr <- df %>%
  filter(!is.na(queue_yr)) %>%
  group_by(queue_yr) %>%
  summarise(
    n_entered_that_yr = n(),
    .groups = "drop"
  ) %>%
  arrange(queue_yr) %>%
  mutate(
    # Cumulative projects entered through end of prior year = backlog proxy
    cum_entered_prior = lag(cumsum(n_entered_that_yr), default = 0),
    # Subtract approximate resolutions: use cumulative completed + withdrawn
  )

# Join resolved counts to estimate active queue size
resolved_by_yr <- df %>%
  filter(!is.na(queue_yr), status_group %in% c("completed", "withdrawn")) %>%
  group_by(queue_yr) %>%
  summarise(n_resolved = n(), .groups = "drop")

congestion_tbl <- congestion_by_yr %>%
  left_join(resolved_by_yr, by = "queue_yr") %>%
  replace_na(list(n_resolved = 0)) %>%
  mutate(
    cum_resolved_prior = lag(cumsum(n_resolved), default = 0),
    # Active backlog at start of year = cumulative entered - cumulative resolved
    queue_backlog      = pmax(cum_entered_prior - cum_resolved_prior, 0),
    # Standardise to mean 0, sd 1 for model interpretability
    congestion_std     = as.numeric(scale(queue_backlog))
  ) %>%
  select(queue_yr, queue_backlog, congestion_std)

model_df_main <- model_df_main %>%
  left_join(congestion_tbl, by = "queue_yr")
```

```{r fig-congestion, echo=FALSE, fig.cap="Queue backlog (active projects) at time of entry by cohort year. Sharp growth post-2005 reflects the renewable energy boom. This is the primary policy lever for queue reform analysis.", fig.height=4}
congestion_tbl %>%
  filter(queue_yr >= 2000, queue_yr <= 2022) %>%
  ggplot(aes(x = queue_yr, y = queue_backlog)) +
  geom_col(fill = "#0077b6", width = 0.8) +
  scale_y_continuous(labels = comma) +
  labs(title   = "Queue Backlog at Time of Entry",
       subtitle = "Number of projects active in queue when a given cohort filed",
       x = "Queue entry year", y = "Active backlog (projects)") +
  theme_minimal(base_size = 12)
```

## 2.2  ISO Reform Adoption

Major ISO queue reforms changed study procedures and cluster processing rules
at different times. We code a `post_reform` binary for each ISO based on known
reform adoption years:

| ISO      | Reform event                        | Year |
|----------|-------------------------------------|------|
| MISO     | Cluster study introduction          | 2008 |
| PJM      | FERC Order 2003 full implementation | 2012 |
| CAISO    | Interconnection process reform      | 2014 |
| All ISOs | FERC Order 2023 (Large Generator)   | 2023 |

```{r lever-reform}
iso_reform_yr <- tibble(
  iso          = c("MISO", "PJM", "CAISO"),
  reform_yr_iso = c(2008L,  2012L,  2014L)
)

model_df_main <- model_df_main %>%
  left_join(iso_reform_yr, by = "iso") %>%
  mutate(
    # ISO-specific reform: 1 if project entered after that ISO's reform year
    post_reform_iso  = as.integer(!is.na(reform_yr_iso) & queue_yr >= reform_yr_iso),
    # FERC Order 2023 applies universally (will only affect very recent cohorts)
    post_order2023   = as.integer(queue_yr >= 2023L),
    # Combined: either ISO reform or Order 2023 in effect
    post_reform      = as.integer(post_reform_iso == 1 | post_order2023 == 1)
  )
```

## 2.3  Technology Era Interactions

Battery storage entered the queue in volume after 2015, coinciding with new
paired-resource interconnection study requirements. Solar and Wind boom-bust
cycles create era effects. We construct a `tech_era` variable to interact with
technology type.

```{r lever-tech-era}
model_df_main <- model_df_main %>%
  mutate(
    era = case_when(
      queue_yr < 2008              ~ "pre-2008",
      queue_yr >= 2008 & queue_yr < 2015 ~ "2008-2014",
      queue_yr >= 2015             ~ "2015+"
    ),
    # REPLACED: Removed "pre-2008" so "2008-2014" becomes the reference level
    era = factor(era, levels = c("2008-2014", "2015+")),

    # Green technology flag — primary policy interest
    is_green = as.integer(tech %in% c("Solar", "Wind", "Battery", "Offshore Wind"))
  )
```

```{r lever-summary, echo=FALSE}
model_df_main %>%
  count(era, iso, .drop = FALSE) %>%
  group_by(era) %>%
  summarise(
    ISOs    = n_distinct(iso),
    Projects = sum(n),
    .groups = "drop"
  ) %>%
  kbl(booktabs = TRUE, caption = "Project counts by technology era") %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE)
```

---

# 3  Competing-Risks Model

Projects face two competing failure events: coming online (event 1) and
withdrawing (event 2). Standard survival analysis treats one as censoring the
other, which is incorrect — a project that withdraws is not "censored" for
coming online, it has permanently exited. Competing-risks models handle both
simultaneously.

**Two complementary approaches:**

- **Cause-specific Cox hazard models** — one model per event type. Coefficients
  estimate the effect of covariates on the hazard of each specific event,
  conditioning on not having experienced either event. Best for understanding
  the mechanisms driving each outcome.
- **Fine-Gray subdistribution hazard model** — directly models the cumulative
  incidence function (CIF) for each event. Accounts for the competing risk
  when estimating the probability of each outcome. Best for policy predictions
  (e.g., "what fraction of projects will come online within 5 years?").

## 3.1  Cause-Specific Hazard: Completion

```{r cs-completion}
# Event 1 indicator: came online; withdrew is treated as censored here
# because we are modelling the hazard of completion specifically.
cs_df <- model_df_main %>%
  mutate(
    event_completion = as.integer(event_code == 1),
    event_withdrawal = as.integer(event_code == 2)
  )

cox_completion <- coxph(
  Surv(tiq_days, event_completion) ~
    congestion_std + post_reform + is_green +
    congestion_std:is_green +
    post_reform:is_green +
    log_mw + iso + era,
  data    = cs_df,
  weights = ipw_trim,
  ties    = "efron"
)
```

```{r cs-completion-table, echo=FALSE}
tidy(cox_completion, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(!str_starts(term, "iso"), !str_starts(term, "era")) %>%
  mutate(
    HR        = round(estimate,  3),
    `95% CI`  = paste0("[", round(conf.low, 3), ", ", round(conf.high, 3), "]"),
    `p-value` = round(p.value, 4),
    Sig       = case_when(p.value < 0.001 ~ "***",
                          p.value < 0.01  ~ "**",
                          p.value < 0.05  ~ "*",
                          TRUE            ~ "")
  ) %>%
  select(Term = term, HR, `95% CI`, `p-value`, Sig) %>%
  kbl(booktabs = TRUE,
      caption = "Cause-specific Cox: completion hazard. HR > 1 = faster to come online.") %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

## 3.2  Cause-Specific Hazard: Withdrawal

```{r cs-withdrawal}
cox_withdrawal <- coxph(
  Surv(tiq_days, event_withdrawal) ~
    congestion_std + post_reform + is_green +
    congestion_std:is_green +
    post_reform:is_green +
    log_mw + iso + era,
  data    = cs_df,
  weights = ipw_trim,
  ties    = "efron"
)
```

```{r cs-withdrawal-table, echo=FALSE}
tidy(cox_withdrawal, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(!str_starts(term, "iso"), !str_starts(term, "era")) %>%
  mutate(
    HR        = round(estimate,  3),
    `95% CI`  = paste0("[", round(conf.low, 3), ", ", round(conf.high, 3), "]"),
    `p-value` = round(p.value, 4),
    Sig       = case_when(p.value < 0.001 ~ "***",
                          p.value < 0.01  ~ "**",
                          p.value < 0.05  ~ "*",
                          TRUE            ~ "")
  ) %>%
  select(Term = term, HR, `95% CI`, `p-value`, Sig) %>%
  kbl(booktabs = TRUE,
      caption = "Cause-specific Cox: withdrawal hazard. HR > 1 = faster to withdraw.") %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

## 3.3  Subdistribution Hazard (Fine-Gray)

The Fine-Gray model gives the subdistribution hazard, which directly maps to
the cumulative incidence function — the probability of experiencing each event
by time t in the presence of competing risks.

```{r fine-gray}
# Convert numeric event_code to a factor for the survival package
cs_df <- cs_df %>%
  mutate(event_factor = factor(event_code, levels = c(0, 1, 2), 
                               labels = c("censored", "online", "withdrew")))

# ── Completion (event = "online") ─────────────────────────────────────────────
fgdata_online <- finegray(
  Surv(tiq_days, event_factor) ~ congestion_std + post_reform + is_green +
    log_mw + iso + era,
  data    = cs_df,
  weights = ipw_trim,      
  etype   = "online"       # Now matches the factor label
)

fg_completion <- tryCatch(
  coxph(
    Surv(fgstart, fgstop, fgstatus) ~
      congestion_std + post_reform + is_green +
      congestion_std:is_green + post_reform:is_green +
      log_mw + iso + era,
    data    = fgdata_online,
    weights = fgwt         
  ),
  error = function(e) { message("Fine-Gray completion failed: ", e$message); NULL }
)

# ── Withdrawal (event = "withdrew") ───────────────────────────────────────────
fgdata_withdrew <- finegray(
  Surv(tiq_days, event_factor) ~ congestion_std + post_reform + is_green +
    log_mw + iso + era,
  data    = cs_df,
  weights = ipw_trim,      
  etype   = "withdrew"     # Now matches the factor label
)

fg_withdrawal <- tryCatch(
  coxph(
    Surv(fgstart, fgstop, fgstatus) ~
      congestion_std + post_reform + is_green +
      congestion_std:is_green + post_reform:is_green +
      log_mw + iso + era,
    data    = fgdata_withdrew,
    weights = fgwt         
  ),
  error = function(e) { message("Fine-Gray withdrawal failed: ", e$message); NULL }
)
```

```{r fine-gray-table, echo=FALSE}
# Helper: tidy a coxph Fine-Gray fit into the summary table format,
# or return an informative placeholder row if the model failed.
format_fg <- function(fit, label) {
  if (is.null(fit)) {
    return(tibble(Model = label, Term = "Model failed — see console",
                  SHR = NA_real_, `p-value` = NA_real_))
  }
  tidy(fit, exponentiate = TRUE, conf.int = FALSE) %>%
    transmute(
      Model     = label,
      Term      = term,
      SHR       = round(estimate, 3),
      `p-value` = round(p.value, 4)
    )
}

bind_rows(
  format_fg(fg_completion, "Completion (event 1)"),
  format_fg(fg_withdrawal, "Withdrawal (event 2)")
) %>%
  filter(!str_starts(Term, "iso"), !str_starts(Term, "era")) %>%
  select(Model, Term, SHR, `p-value`) %>%
  kbl(booktabs = TRUE,
      caption = "Fine-Gray subdistribution hazard ratios (via finegray + coxph). SHR > 1 = higher CIF for that event.") %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

## 3.4  Cause-Specific Hazard Ratios — Visual Comparison

```{r fig-cs-comparison, echo=FALSE, fig.cap="Cause-specific hazard ratios for policy levers and green technology interactions. Bars to the right of 1.0 (completion) indicate faster completion; bars to the right of 1.0 (withdrawal) indicate faster exit.", fig.height=5}
tidy_completion <- tidy(cox_completion, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(!str_starts(term, "iso"), !str_starts(term, "era")) %>%
  mutate(model = "Completion hazard")

tidy_withdrawal <- tidy(cox_withdrawal, exponentiate = TRUE, conf.int = TRUE) %>%
  filter(!str_starts(term, "iso"), !str_starts(term, "era")) %>%
  mutate(model = "Withdrawal hazard")

bind_rows(tidy_completion, tidy_withdrawal) %>%
  mutate(
    model = factor(model, levels = c("Completion hazard", "Withdrawal hazard")),
    term  = str_replace_all(term, "_", " ")
  ) %>%
  ggplot(aes(x = estimate, y = reorder(term, estimate),
             colour = model, shape = model)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c("Completion hazard" = "#0077b6",
                                 "Withdrawal hazard"  = "#e63946")) +
  facet_wrap(~model, scales = "free_x") +
  labs(title  = "Cause-Specific Hazard Ratios — Policy Levers",
       x = "Hazard Ratio (95% CI)", y = NULL,
       colour = NULL, shape = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

---

# 4  Parametric AFT Model (Time to Online)

The Cox model gives relative hazards but no absolute time predictions. A
parametric accelerated failure time (AFT) model provides interpretable
predictions: "under this policy lever, projects come online X days faster."
We restrict to **completion events only** and treat withdrawals as censored.

**Distribution selection:** Log-normal is preferred for interconnection queue
times — the distribution is right-skewed with a long tail (some projects take
10+ years) but not monotonically hazardous.

```{r aft-fit}
aft_df <- cs_df %>%
  filter(event_code %in% c(0L, 1L)) %>%
  droplevels()   # remove empty factor levels (e.g. era "pre-2008" has zero rows
                 # because model_df_main is already filtered to queue_yr >= 2008;
                 # any NA coefficient from an empty level propagates to all predictions)
# event_code 2 (withdrew) is right-censored here — we observe they left but
# not when they would have come online had they stayed.

aft_lognormal <- survreg(
  Surv(tiq_days, event_completion) ~
    congestion_std + post_reform + is_green +
    congestion_std:is_green +
    post_reform:is_green +
    log_mw + iso + era,
  data    = aft_df,
  weights = ipw_trim,
  dist    = "lognormal"
)

aft_weibull <- survreg(
  Surv(tiq_days, event_completion) ~
    congestion_std + post_reform + is_green +
    congestion_std:is_green +
    post_reform:is_green +
    log_mw + iso + era,
  data    = aft_df,
  weights = ipw_trim,
  dist    = "weibull"
)
```

```{r aft-model-selection, echo=FALSE}
tibble(
  Distribution = c("Log-normal", "Weibull"),
  LogLik       = c(round(logLik(aft_lognormal), 1),
                   round(logLik(aft_weibull),   1)),
  AIC          = c(round(AIC(aft_lognormal), 1),
                   round(AIC(aft_weibull),   1))
) %>%
  kbl(booktabs = TRUE, caption = "AFT model fit comparison. Lower AIC = better fit.") %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

```{r aft-table, echo=FALSE}
# AFT coefficients: exp(coef) is the time ratio — values < 1 mean faster
tidy(aft_lognormal, conf.int = TRUE) %>%
  filter(!str_starts(term, "iso"), !str_starts(term, "era"),
         term != "Log(scale)") %>%
  mutate(
    `Time ratio`  = round(exp(estimate), 3),
    `95% CI`      = paste0("[", round(exp(conf.low), 3), ", ",
                            round(exp(conf.high), 3), "]"),
    `p-value`     = round(p.value, 4),
    Interpretation = case_when(
      exp(estimate) < 1 ~ paste0(round((1 - exp(estimate)) * 100, 1), "% faster"),
      exp(estimate) > 1 ~ paste0(round((exp(estimate) - 1) * 100, 1), "% slower"),
      TRUE              ~ "no effect"
    ),
    Sig = case_when(p.value < 0.001 ~ "***",
                    p.value < 0.01  ~ "**",
                    p.value < 0.05  ~ "*",
                    TRUE            ~ "")
  ) %>%
  select(Term = term, `Time ratio`, `95% CI`, `p-value`, Sig, Interpretation) %>%
  kbl(booktabs = TRUE,
      caption = "Log-normal AFT model (time to online). Time ratio < 1 = faster completion.") %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

## 4.1  Predicted Completion Time Under Policy Scenarios

Using the log-normal AFT, we predict median completion time for a reference
project under three scenarios:

- **Baseline:** mean congestion, no ISO reform, green technology (Solar 100 MW, PJM, 2008–2014 era)
- **Reform:** same as baseline but `post_reform = 1`
- **Low congestion:** congestion reduced by 1 SD (−1 from mean)

```{r aft-scenarios}
# iso and era passed as strings — predict.survreg coerces to match training
# factor levels automatically, avoiding factor/level mismatch issues.
ref_project <- tibble(
  congestion_std = c(0,  0, -1),
  post_reform    = c(0L, 1L, 0L),
  is_green       = c(1L, 1L, 1L),
  log_mw         = log1p(100),
  iso            = "PJM",
  era            = "2008-2014",
  Scenario       = c("Baseline", "Post-reform", "Low congestion (−1 SD)")
)

# predict.survreg with type = "quantile" returns a matrix; as.numeric() flattens
# it to a plain vector so dplyr and kbl handle it correctly.
ref_project <- ref_project %>%
  mutate(
    median_days  = as.numeric(predict(aft_lognormal, newdata = as.data.frame(ref_project),
                                      type = "quantile", p = 0.5)),
    median_years = round(median_days / 365.25, 2)
  )
```

```{r aft-scenario-table, echo=FALSE}
ref_project %>%
  select(Scenario, `Median TIQ (days)` = median_days,
         `Median TIQ (years)` = median_years) %>%
  mutate(`Median TIQ (days)` = round(`Median TIQ (days)`, 0)) %>%
  kbl(booktabs = TRUE,
      caption = "Predicted median time-to-online under three policy scenarios. Reference: Solar 100 MW, PJM, 2008–2014 era.") %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```


---

# 5  Capacity-Weighted Outcomes (MW Objective)

The MW objective requires weighting predicted outcomes by project capacity.
A policy that accelerates 10 × 100 MW Solar projects is more valuable than
one that accelerates 100 × 10 MW Gas projects.

## 5.1  Predicted Completion Probability by Technology

Using the Fine-Gray completion model, we predict the 5-year cumulative
incidence of completion for each technology at mean congestion, post-reform.

```{r mw-setup}
# For MW-weighted analysis, use the cause-specific completion hazard
# and predict at fixed covariate values varying only by technology.

tech_levels <- levels(cs_df$tech)

tech_scenarios <- tibble(
  tech           = factor(tech_levels, levels = tech_levels),
  congestion_std = 0,
  post_reform    = 1L,
  is_green       = as.integer(tech_levels %in% c("Solar", "Wind", "Battery", "Offshore Wind")),
  log_mw         = log1p(100),
  iso            = "PJM",    # FIXED: Passed as string to avoid factor NA bugs
  era            = "2015+"   # FIXED: Passed as string to avoid factor NA bugs
)
```

## 5.2  MW-Weighted Scenario Analysis

For each technology, we compute: expected MW online within 5 years =
predicted completion probability × nameplate MW in queue.

```{r mw-weighted}
# Mean MW by technology (in full dataset)
mw_by_tech <- model_df_main %>%
  filter(!is.na(mw_num)) %>%
  group_by(tech) %>%
  summarise(
    total_mw   = sum(mw_num, na.rm = TRUE),
    n_projects = n(),
    mean_mw    = mean(mw_num, na.rm = TRUE),
    .groups    = "drop"
  )

# Predicted median completion time from AFT (type = "quantile", p = 0.5)
# Returns median in days directly; derive log_median_days for the pnorm step.
tech_scenarios <- tech_scenarios %>%
  mutate(
    # FIXED: Added as.numeric() to flatten the matrix returned by predict!
    # Without this, predict returns a 2D matrix which dplyr cannot display in a table.
    median_days     = as.numeric(predict(aft_lognormal, newdata = as.data.frame(tech_scenarios),
                                         type = "quantile", p = 0.5)),
    log_median_days = log(median_days),   # log-median = Xβ; needed for pnorm CDF
    median_years    = round(median_days / 365.25, 2)
  ) %>%
  left_join(mw_by_tech, by = "tech") %>%
  mutate(
    # P(T < 5yr | log-normal) = pnorm((log(5*365.25) - log_median_days) / aft_lognormal$scale)
    p_complete_5yr  = pnorm(
      (log(5 * 365.25) - log_median_days) / aft_lognormal$scale
    ),
    mw_expected_5yr = p_complete_5yr * total_mw / 1e3  # GW
  )
```

```{r mw-table, echo=FALSE}
tech_scenarios %>%
  arrange(desc(mw_expected_5yr)) %>%
  mutate(
    is_green    = ifelse(is_green == 1, "Yes", "No"),
    median_years = round(median_years, 2),
    p_complete_5yr = paste0(round(100 * p_complete_5yr, 1), "%"),
    mw_expected_5yr = round(mw_expected_5yr, 1),
    total_mw    = round(total_mw / 1e3, 1)
  ) %>%
  select(Technology = tech, Green = is_green,
         `Median TIQ (yr)` = median_years,
         `P(complete ≤5yr)` = p_complete_5yr,
         `Total MW in queue (GW)` = total_mw,
         `Expected GW online ≤5yr` = mw_expected_5yr) %>%
  kbl(booktabs = TRUE,
      caption = "Capacity-weighted completion forecast (post-reform, PJM reference, 2015+ era). Expected GW online = P(complete ≤5yr) × total queue MW.") %>%
  kable_styling(latex_options = c("striped", "hold_position", "scale_down"),
                full_width = FALSE) %>%
  row_spec(which(tech_scenarios %>% arrange(desc(mw_expected_5yr)) %>% pull(is_green) == 1),
           bold = TRUE) %>%
  row_spec(0, bold = TRUE)
```

```{r fig-mw-outcomes, echo=FALSE, fig.cap="Expected GW online within 5 years by technology (post-reform scenario). Bold = green technology. Height reflects both completion probability and total MW in queue.", fig.height=4}
tech_scenarios %>%
  arrange(desc(mw_expected_5yr)) %>%
  mutate(
    tech   = factor(tech, levels = rev(tech)),
    colour = ifelse(is_green == 1, "#43AA8B", "#adb5bd")
  ) %>%
  ggplot(aes(x = mw_expected_5yr, y = tech, fill = factor(is_green))) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c("0" = "#adb5bd", "1" = "#43AA8B"),
                    labels = c("0" = "Conventional", "1" = "Green"),
                    name = NULL) +
  scale_x_continuous(labels = comma) +
  labs(title    = "Expected GW Online Within 5 Years by Technology",
       subtitle = "Post-reform scenario, PJM reference, 2015+ era. Green = Solar/Wind/Battery/Offshore.",
       x = "Expected GW online (≤5yr)", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

---

# 6  Summary

```{r summary-table, echo=FALSE}
# Pull key coefficients for the summary
hr_congestion_completion <- round(exp(coef(cox_completion)["congestion_std"]), 3)
hr_reform_completion     <- round(exp(coef(cox_completion)["post_reform"]), 3)
hr_green_completion      <- round(exp(coef(cox_completion)["is_green"]), 3)

hr_congestion_withdrawal <- round(exp(coef(cox_withdrawal)["congestion_std"]), 3)
hr_reform_withdrawal     <- round(exp(coef(cox_withdrawal)["post_reform"]), 3)
hr_green_withdrawal      <- round(exp(coef(cox_withdrawal)["is_green"]), 3)

tr_congestion <- round(exp(coef(aft_lognormal)["congestion_std"]), 3)
tr_reform     <- round(exp(coef(aft_lognormal)["post_reform"]), 3)
tr_green      <- round(exp(coef(aft_lognormal)["is_green"]), 3)

tibble(
  Lever = c("Queue congestion (+1 SD)",
            "Post ISO reform",
            "Green technology"),
  `Completion HR` = c(hr_congestion_completion, hr_reform_completion, hr_green_completion),
  `Withdrawal HR` = c(hr_congestion_withdrawal, hr_reform_withdrawal, hr_green_withdrawal),
  `AFT time ratio` = c(tr_congestion, tr_reform, tr_green)
) %>%
  kbl(booktabs = TRUE,
      caption = "Key policy lever estimates across all three model frameworks. HR > 1 = higher hazard; time ratio < 1 = faster completion.") %>%
  kable_styling(latex_options = c("striped", "hold_position"), full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

**Scope and confidence notes:**

- All models include ISO and era fixed effects. Estimates are identified from
  within-ISO, within-era variation in policy levers and project characteristics.
- SPP and MISO carry lower confidence due to EIA-860 coverage limitations
  (5% and 28% match rates). Excluding SPP and MISO and rerunning is a
  recommended robustness check.
- The congestion lever is a cohort-year approximation. A more precise
  measure would use project-level entry dates to count active queue members
  at the exact moment of filing.
- Residual confounding from unobserved factors (developer resources, local
  permitting, transmission congestion severity) cannot be ruled out.

---

*Generated: `r Sys.Date()` | Data: `ISO_EIA_Merged_v7.csv` | Stage 2 of PlanningQueues policy model pipeline*
