---
title: "Interconnection Queue Time by Generation Type (2000–2019)"
subtitle: "How queue durations differ and have changed across technology types"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: hide
    fig_width: 12
    fig_height: 7
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")

pkgs <- c("tidyverse","survival","survminer","quantreg","Kendall",
          "scales","patchwork","ggridges","knitr","kableExtra","viridis","ggrepel")
new  <- pkgs[!pkgs %in% installed.packages()[,"Package"]]
if (length(new)) install.packages(new, repos = "https://cloud.r-project.org", quiet = TRUE)

library(tidyverse); library(survival); library(survminer)
library(quantreg);  library(Kendall);  library(scales)
library(patchwork); library(ggridges); library(knitr)
library(kableExtra);library(viridis);  library(ggrepel)
```

```{r load-data}
DATA_DIR <- Sys.getenv("PLANNING_QUEUES_DATA", "../../data")
df_raw <- read_csv(
  file.path(DATA_DIR, "TIQ_Analysis_Clean.csv"),
  show_col_types = FALSE
)

df <- df_raw %>%
  mutate(
    tiq_days  = as.numeric(time_in_queue_days),
    tiq_years = tiq_days / 365.25,
    entry_year = as.integer(queue_entry_year),
    mw        = as.numeric(mw1),
    outcome   = factor(outcome, levels = c("came_online","withdrew")),
    tech      = factor(type1),
    era       = case_when(
      entry_year %in% 2000:2004 ~ "2000–2004",
      entry_year %in% 2005:2009 ~ "2005–2009",
      entry_year %in% 2010:2014 ~ "2010–2014",
      entry_year %in% 2015:2019 ~ "2015–2019"
    ),
    era    = factor(era, levels = c("2000–2004","2005–2009","2010–2014","2015–2019")),
    decade = ifelse(entry_year <= 2009, "2000–2009", "2010–2019")
  ) %>%
  filter(!is.na(tiq_days), !is.na(entry_year))

# Focus techs — sufficient data for trend analysis
FOCUS_TECHS    <- c("Solar","Wind","Gas","Battery","Coal","Nuclear","Hydro","Offshore Wind")
TREND_TECHS    <- c("Solar","Wind","Gas","Battery")   # 20 usable years
SECONDARY_TECHS <- c("Coal","Hydro","Nuclear","Offshore Wind")  # fewer years but meaningful

df_focus <- df %>% filter(tech %in% FOCUS_TECHS) %>%
  mutate(tech = factor(tech, levels = FOCUS_TECHS))

co <- df_focus %>% filter(outcome == "came_online")
wd <- df_focus %>% filter(outcome == "withdrew")

# Palettes
pal_tech <- c(
  "Solar"         = "#F4A261",
  "Wind"          = "#43AA8B",
  "Gas"           = "#E63946",
  "Battery"       = "#577590",
  "Coal"          = "#6D4C41",
  "Nuclear"       = "#9C27B0",
  "Hydro"         = "#0077B6",
  "Offshore Wind" = "#00B4D8"
)
pal_outcome <- c("came_online" = "#2196F3", "withdrew" = "#FF6B35")
pal_era     <- c("2000–2004"="#8B0000","2005–2009"="#E85D04",
                 "2010–2014"="#F4A261","2015–2019"="#2A9D8F")

theme_queue <- theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 14, margin = margin(b = 5)),
    plot.subtitle = element_text(colour = "grey40", size = 11, margin = margin(b = 8)),
    plot.caption  = element_text(colour = "grey55", size = 9, hjust = 0),
    legend.position  = "bottom",
    panel.grid.minor = element_blank(),
    strip.text    = element_text(face = "bold", size = 11)
  )
theme_set(theme_queue)
```

---

# 1 · Technology Overview

```{r tech-overview}
tech_summary <- df_focus %>%
  group_by(tech, outcome) %>%
  summarise(
    n          = n(),
    med_days   = round(median(tiq_days)),
    med_years  = round(median(tiq_days) / 365.25, 1),
    p25        = round(quantile(tiq_days, .25)),
    p75        = round(quantile(tiq_days, .75)),
    mean_mw    = round(mean(mw, na.rm = TRUE)),
    .groups    = "drop"
  )

tech_summary %>%
  mutate(outcome = recode(outcome,
    "came_online" = "Came Online", "withdrew" = "Withdrew")) %>%
  arrange(tech, outcome) %>%
  kbl(
    caption   = "Queue time summary by technology and outcome",
    col.names = c("Technology","Outcome","N","Median (days)","Median (years)",
                  "P25 (days)","P75 (days)","Mean MW")
  ) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE) %>%
  collapse_rows(columns = 1, valign = "middle") %>%
  row_spec(0, bold = TRUE)
```

```{r tech-count-bar, fig.height=5}
df_focus %>%
  count(tech, outcome) %>%
  ggplot(aes(x = reorder(tech, n), y = n, fill = outcome)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(aes(label = comma(n)),
            position = position_stack(vjust = 0.5),
            colour = "white", fontface = "bold", size = 3.2) +
  coord_flip() +
  scale_fill_manual(values = pal_outcome,
                    labels = c("Came Online","Withdrew")) +
  scale_y_continuous(labels = comma) +
  labs(title = "Queue Filings by Technology and Outcome",
       subtitle = "Solar and Wind dominate; Battery is entirely post-2008",
       x = NULL, y = "Number of projects", fill = NULL,
       caption = "Cohorts 2000–2019 with computable queue duration")
```

```{r tech-median-bar, fig.height=5}
tech_summary %>%
  mutate(outcome_lab = recode(outcome,
    "came_online" = "Came Online", "withdrew" = "Withdrew")) %>%
  ggplot(aes(x = reorder(tech, med_days), y = med_days / 365.25,
             fill = tech, alpha = outcome)) +
  geom_col(position = position_dodge(0.75), width = 0.65) +
  geom_text(aes(label = paste0(round(med_days / 365.25, 1),"y")),
            position = position_dodge(0.75), hjust = -0.1,
            size = 3, fontface = "bold") +
  coord_flip() +
  scale_fill_manual(values = pal_tech, guide = "none") +
  scale_alpha_manual(values = c("came_online" = 1, "withdrew" = 0.45),
                     labels = c("Came Online","Withdrew"), name = NULL) +
  scale_y_continuous(labels = function(x) paste0(x,"y"), limits = c(0, 7)) +
  labs(title = "Median Queue Duration by Technology",
       subtitle = "Solid = came online · Faded = withdrew",
       x = NULL, y = "Median years in queue",
       caption = "Nuclear and Offshore Wind have the longest median durations for successful projects")
```

---

# 2 · Distribution Shape by Technology

```{r ridge-by-tech, fig.height=9}
df_focus %>%
  filter(!is.na(era)) %>%
  ggplot(aes(x = tiq_days / 365.25,
             y = fct_rev(factor(tech, levels = FOCUS_TECHS)),
             fill = tech)) +
  geom_density_ridges(
    aes(height = after_stat(density)),
    alpha = 0.8, scale = 1.6, rel_min_height = 0.01,
    quantile_lines = TRUE, quantiles = c(0.25, 0.5, 0.75)
  ) +
  scale_fill_manual(values = pal_tech, guide = "none") +
  scale_x_continuous(labels = function(x) paste0(x,"y"),
                     limits = c(0, 15), breaks = seq(0, 15, 3)) +
  facet_wrap(~ outcome,
             labeller = labeller(outcome = c(
               "came_online" = "Came Online",
               "withdrew"    = "Withdrew"))) +
  labs(title = "Full Distribution of Queue Times by Technology",
       subtitle = "Vertical lines = P25, P50, P75. Each row is one technology.",
       x = "Years in queue", y = NULL,
       caption = "Outliers above 15 years excluded for readability")
```

---

# 3 · Trend Over Time by Technology

## 3.1 Annual Median Trends

```{r annual-trend-facet, fig.height=10}
annual_tech <- df_focus %>%
  group_by(tech, outcome, entry_year) %>%
  summarise(
    med = median(tiq_days),
    n   = n(),
    q25 = quantile(tiq_days, 0.25),
    q75 = quantile(tiq_days, 0.75),
    .groups = "drop"
  ) %>%
  filter(n >= 5)

ggplot(annual_tech, aes(x = entry_year, y = med / 365.25,
                         colour = outcome, group = outcome)) +
  geom_ribbon(aes(ymin = q25 / 365.25, ymax = q75 / 365.25,
                  fill = outcome, colour = NULL),
              alpha = 0.12) +
  geom_line(linewidth = 1) +
  geom_point(aes(size = n), alpha = 0.75) +
  geom_smooth(method = "lm", se = FALSE,
              linewidth = 0.5, linetype = "dashed") +
  facet_wrap(~ tech, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = pal_outcome,
                      labels = c("Came Online","Withdrew")) +
  scale_fill_manual(values   = pal_outcome, guide = "none") +
  scale_size_continuous(range = c(1, 5), guide = "none") +
  scale_x_continuous(breaks = seq(2000, 2019, 5)) +
  scale_y_continuous(labels = function(x) paste0(x,"y")) +
  labs(title = "Annual Median Queue Time by Technology and Outcome",
       subtitle = "Ribbon = IQR. Dashed line = OLS trend. Free y-axis scales.",
       x = "Queue entry year", y = "Median years in queue",
       colour = NULL,
       caption = "Points shown only where n ≥ 5 for that technology × year × outcome")
```

## 3.2 Era Comparison: 2000s vs 2010s

```{r era-by-tech, fig.height=8}
era_tech <- df_focus %>%
  filter(!is.na(decade)) %>%
  group_by(tech, outcome, decade) %>%
  summarise(
    med = median(tiq_days),
    n   = n(),
    .groups = "drop"
  )

# compute % change
era_wide <- era_tech %>%
  pivot_wider(names_from = decade, values_from = c(med, n)) %>%
  rename(med_early = `med_2000–2009`, med_late = `med_2010–2019`,
         n_early   = `n_2000–2009`,   n_late   = `n_2010–2019`) %>%
  mutate(pct_change = 100 * (med_late - med_early) / med_early) %>%
  filter(!is.na(med_early), !is.na(med_late), n_early >= 10, n_late >= 10)

ggplot(era_wide,
       aes(x = med_early / 365.25, xend = med_late / 365.25,
           y = reorder(tech, med_early),
           colour = outcome)) +
  geom_segment(linewidth = 2.5, alpha = 0.8,
               arrow = arrow(length = unit(0.25,"cm"), type = "closed")) +
  geom_point(aes(x = med_early / 365.25), size = 3.5, shape = 21,
             fill = "white", stroke = 1.5) +
  geom_text(aes(x = (med_early + med_late) / 2 / 365.25,
                label = sprintf("%+.0f%%", pct_change)),
            vjust = -0.8, size = 3.2, fontface = "bold") +
  facet_wrap(~ outcome,
             labeller = labeller(outcome = c(
               "came_online" = "Came Online",
               "withdrew"    = "Withdrew"))) +
  scale_colour_manual(values = pal_outcome, guide = "none") +
  scale_x_continuous(labels = function(x) paste0(x,"y")) +
  labs(title = "Shift in Median Queue Time: 2000s → 2010s by Technology",
       subtitle = "Circle = 2000s median. Arrowhead = 2010s median. % = relative change.",
       x = "Median years in queue", y = NULL,
       caption = "Only shown where n ≥ 10 in both decades")
```

---

# 4 · Mann-Kendall Trend Tests by Technology

```{r mk-by-tech}
mk_rows <- list()

for (tech in FOCUS_TECHS) {
  for (out in c("came_online","withdrew")) {
    sub <- df_focus %>%
      filter(type1 == tech, outcome == out) %>%
      group_by(entry_year) %>%
      summarise(med = median(tiq_days), n = n(), .groups = "drop") %>%
      filter(n >= 5) %>%
      arrange(entry_year)

    if (nrow(sub) < 6) {
      mk_rows[[paste(tech, out)]] <- tibble(
        Technology  = tech,
        Outcome     = ifelse(out == "came_online","Came Online","Withdrew"),
        N_years     = nrow(sub),
        Tau         = NA_real_,
        P_value     = NA_real_,
        Direction   = "Insufficient data",
        Significance= "—"
      )
      next
    }

    mk       <- MannKendall(sub$med)
    mk_rows[[paste(tech, out)]] <- tibble(
      Technology   = tech,
      Outcome      = ifelse(out == "came_online","Came Online","Withdrew"),
      N_years      = nrow(sub),
      Tau          = round(mk$tau[1], 3),
      P_value      = round(mk$sl[1], 4),
      Direction    = case_when(
        mk$tau[1] >  0.1 ~ "↑ Increasing",
        mk$tau[1] < -0.1 ~ "↓ Decreasing",
        TRUE              ~ "→ Flat"
      ),
      Significance = case_when(
        mk$sl[1] < 0.01 ~ "*** p<0.01",
        mk$sl[1] < 0.05 ~ "**  p<0.05",
        mk$sl[1] < 0.10 ~ "*   p<0.10",
        TRUE             ~ "ns"
      )
    )
  }
}

mk_tech <- bind_rows(mk_rows)

mk_tech %>%
  kbl(caption = "Mann-Kendall trend tests by technology and outcome",
      col.names = c("Technology","Outcome","Years","Tau","P-value",
                    "Direction","Significance")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE) %>%
  row_spec(which(!is.na(mk_tech$P_value) & mk_tech$P_value < 0.05),
           background = "#f0f8ff") %>%
  column_spec(6, bold = TRUE,
              color = case_when(
                grepl("↑", mk_tech$Direction) ~ "#1565C0",
                grepl("↓", mk_tech$Direction) ~ "#B71C1C",
                TRUE                          ~ "grey50"
              )) %>%
  collapse_rows(columns = 1, valign = "middle")
```

```{r mk-heatmap, fig.height=4.5}
mk_tech %>%
  filter(!is.na(Tau)) %>%
  mutate(
    sig_label = case_when(
      P_value < 0.01 ~ "***",
      P_value < 0.05 ~ "**",
      P_value < 0.10 ~ "*",
      TRUE           ~ ""
    ),
    cell_label = paste0(round(Tau, 2), sig_label)
  ) %>%
  ggplot(aes(x = Outcome,
             y = factor(Technology, levels = rev(FOCUS_TECHS)),
             fill = Tau)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  geom_text(aes(label = cell_label), fontface = "bold", size = 4) +
  scale_fill_gradient2(
    low      = "#B71C1C",
    mid      = "white",
    high     = "#1565C0",
    midpoint = 0,
    limits   = c(-1, 1),
    name     = "Kendall's τ\n(blue = increasing)"
  ) +
  labs(title = "Mann-Kendall τ Heatmap by Technology and Outcome",
       subtitle = "Blue = queue times increasing · Red = decreasing · * p<0.1, ** p<0.05, *** p<0.01",
       x = NULL, y = NULL,
       caption = "Cells without label = insufficient annual data points (< 6 years)")
```

---

# 5 · Kaplan-Meier Survival Curves by Technology

> Each panel shows survival curves for the four cohort eras.
> **Came Online**: downward shift = longer time to completion.
> **Withdrew**: upward shift = developers holding on longer before giving up.

## 5.1 Came Online

```{r km-co-tech, fig.height=14, fig.width=13}
km_plots_co <- list()

for (tech in TREND_TECHS) {
  sub <- co %>%
    filter(type1 == tech, !is.na(era)) %>%
    mutate(event = 1)

  if (nrow(sub) < 20) next

  fit <- survfit(Surv(tiq_days, event) ~ era, data = sub)

  p <- ggsurvplot(
    fit,
    data          = sub,
    palette       = unname(pal_era),
    conf.int      = TRUE,
    conf.int.alpha= 0.10,
    pval          = TRUE,
    pval.coord    = c(3200, 0.92),
    xlab          = "Days in queue",
    ylab          = "P(still in queue)",
    title         = tech,
    legend        = "none",
    xlim          = c(0, 3500),
    break.x.by    = 700,
    ggtheme       = theme_queue + theme(plot.title = element_text(
      colour = pal_tech[tech], size = 13))
  )
  km_plots_co[[tech]] <- p$plot
}

wrap_plots(km_plots_co, ncol = 2) +
  plot_annotation(
    title    = "Kaplan-Meier: Came Online — by Technology and Era",
    subtitle = "Downward shift across eras = longer time to reach commercial operation",
    caption  = "Colours: dark red = 2000–2004 → teal = 2015–2019. Log-rank p-value shown.",
    theme    = theme(plot.title = element_text(face = "bold", size = 15))
  )
```

## 5.2 Withdrew

```{r km-wd-tech, fig.height=14, fig.width=13}
km_plots_wd <- list()

for (tech in TREND_TECHS) {
  sub <- wd %>%
    filter(type1 == tech, !is.na(era)) %>%
    mutate(event = 1)

  if (nrow(sub) < 20) next

  fit <- survfit(Surv(tiq_days, event) ~ era, data = sub)

  p <- ggsurvplot(
    fit,
    data          = sub,
    palette       = unname(pal_era),
    conf.int      = TRUE,
    conf.int.alpha= 0.10,
    pval          = TRUE,
    pval.coord    = c(3200, 0.92),
    xlab          = "Days in queue",
    ylab          = "P(still in queue)",
    title         = tech,
    legend        = "none",
    xlim          = c(0, 3500),
    break.x.by    = 700,
    ggtheme       = theme_queue + theme(plot.title = element_text(
      colour = pal_tech[tech], size = 13))
  )
  km_plots_wd[[tech]] <- p$plot
}

wrap_plots(km_plots_wd, ncol = 2) +
  plot_annotation(
    title    = "Kaplan-Meier: Withdrew — by Technology and Era",
    subtitle = "Upward shift = projects holding on longer before exiting. Downward shift = faster exit.",
    caption  = "Colours: dark red = 2000–2004 → teal = 2015–2019.",
    theme    = theme(plot.title = element_text(face = "bold", size = 15))
  )
```

---

# 6 · Quantile Regression by Technology

> Does the trend live in the middle of the distribution or in the tails?
> Steeper P90 slope vs P50 = the longest projects are getting disproportionately worse.

```{r qr-by-tech, fig.height=10}
quantiles <- c(0.25, 0.50, 0.75, 0.90)
qr_tech_preds <- list()

for (tech in TREND_TECHS) {
  for (out in c("came_online","withdrew")) {
    sub <- df_focus %>%
      filter(type1 == tech, outcome == out, tiq_days > 0)
    if (nrow(sub) < 50) next

    for (q in quantiles) {
      tryCatch({
        fit    <- rq(tiq_days ~ entry_year, tau = q, data = sub)
        s      <- summary(fit, se = "boot", R = 300)
        coefs  <- as.data.frame(s$coefficients)
        slope  <- coefs["entry_year","Value"]
        int    <- coefs["(Intercept)","Value"]

        pred_grid <- tibble(
          entry_year = 2000:2019,
          pred       = int + slope * entry_year,
          quantile   = q,
          label      = paste0("P", q * 100),
          tech       = tech,
          outcome    = out
        )
        qr_tech_preds[[paste(tech, out, q)]] <- pred_grid
      }, error = function(e) NULL)
    }
  }
}

qr_df <- bind_rows(qr_tech_preds) %>%
  mutate(outcome_lab = recode(outcome,
    "came_online" = "Came Online", "withdrew" = "Withdrew"))

ggplot(qr_df %>% filter(tech %in% TREND_TECHS),
       aes(x = entry_year, y = pred / 365.25,
           colour = factor(label, levels = paste0("P", quantiles * 100)),
           group  = label)) +
  geom_line(linewidth = 1.1) +
  geom_label_repel(
    data = qr_df %>%
      filter(tech %in% TREND_TECHS, entry_year == 2019),
    aes(label = label),
    nudge_x = 0.6, size = 2.8, fontface = "bold",
    show.legend = FALSE, min.segment.length = 0.2
  ) +
  facet_grid(tech ~ outcome_lab, scales = "free_y") +
  scale_colour_viridis_d(option = "plasma", name = "Percentile") +
  scale_x_continuous(breaks = seq(2000, 2019, 5)) +
  scale_y_continuous(labels = function(x) paste0(x,"y")) +
  labs(title = "Quantile Regression Fan by Technology and Outcome",
       subtitle = "Steeper P90 vs P50 = tail projects getting disproportionately longer",
       x = "Queue entry year", y = "Predicted queue duration",
       caption = "Fitted via quantile regression with bootstrapped SEs (R=300)")
```

---

# 7 · Decade Comparison by Technology — Statistical Tests

```{r wilcoxon-tech}
wx_rows <- list()

for (tech in FOCUS_TECHS) {
  for (out in c("came_online","withdrew")) {
    sub <- df_focus %>%
      filter(type1 == tech, outcome == out)

    early <- sub %>% filter(entry_year <= 2009) %>% pull(tiq_days)
    late  <- sub %>% filter(entry_year >= 2010) %>% pull(tiq_days)

    if (length(early) < 10 | length(late) < 10) next

    wx <- wilcox.test(early, late, alternative = "two.sided")

    wx_rows[[paste(tech, out)]] <- tibble(
      Technology = tech,
      Outcome    = ifelse(out == "came_online","Came Online","Withdrew"),
      N_2000s    = length(early),
      N_2010s    = length(late),
      Med_2000s  = round(median(early)),
      Med_2010s  = round(median(late)),
      Pct_Change = round(100 * (median(late) - median(early)) / median(early), 1),
      P_value    = signif(wx$p.value, 3),
      Sig        = case_when(
        wx$p.value < 0.001 ~ "***",
        wx$p.value < 0.01  ~ "**",
        wx$p.value < 0.05  ~ "*",
        wx$p.value < 0.10  ~ ".",
        TRUE               ~ "ns"
      )
    )
  }
}

wx_df <- bind_rows(wx_rows)

wx_df %>%
  arrange(Technology, Outcome) %>%
  kbl(caption = "Wilcoxon rank-sum test: 2000s vs 2010s median queue times by technology",
      col.names = c("Technology","Outcome","N 2000s","N 2010s",
                    "Median 2000s (d)","Median 2010s (d)",
                    "% Change","P-value","Sig.")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE) %>%
  column_spec(7, bold = TRUE,
              color = ifelse(wx_df$Pct_Change > 0, "#1565C0", "#B71C1C")) %>%
  row_spec(which(wx_df$P_value < 0.05), background = "#f0f8ff") %>%
  collapse_rows(columns = 1, valign = "middle")
```

---

# 8 · Secondary Technologies

> Coal, Nuclear, Hydro, and Offshore Wind have smaller samples but distinct patterns.

```{r secondary-tech, fig.height=7}
sec_annual <- df %>%
  filter(type1 %in% SECONDARY_TECHS) %>%
  group_by(type1, outcome, entry_year) %>%
  summarise(med = median(tiq_days), n = n(), .groups = "drop") %>%
  filter(n >= 5)

ggplot(sec_annual, aes(x = entry_year, y = med / 365.25,
                        colour = outcome, group = outcome)) +
  geom_line(linewidth = 1) +
  geom_point(aes(size = n), alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, linetype = "dashed") +
  facet_wrap(~ type1, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = pal_outcome,
                      labels = c("Came Online","Withdrew")) +
  scale_size_continuous(range = c(2, 6), guide = "none") +
  scale_x_continuous(breaks = seq(2000, 2019, 5)) +
  scale_y_continuous(labels = function(x) paste0(x, "y")) +
  labs(title = "Queue Time Trends — Secondary Technologies",
       subtitle = "Smaller samples; interpret directionally. Point size ∝ n.",
       x = "Queue entry year", y = "Median years in queue",
       colour = NULL,
       caption = "Dashed = OLS trend. Free y-axis scales.")
```

```{r secondary-ridge, fig.height=6}
df %>%
  filter(type1 %in% SECONDARY_TECHS, !is.na(era)) %>%
  mutate(type1 = factor(type1, levels = SECONDARY_TECHS)) %>%
  ggplot(aes(x = tiq_days / 365.25, y = era, fill = era)) +
  geom_density_ridges(alpha = 0.75, scale = 1.4, rel_min_height = 0.01,
                      quantile_lines = TRUE, quantiles = c(0.25, 0.5, 0.75)) +
  facet_wrap(~ type1, scales = "free_x", ncol = 2) +
  scale_fill_manual(values = pal_era, guide = "none") +
  scale_x_continuous(labels = function(x) paste0(x,"y"), limits = c(0, 20)) +
  labs(title = "Distribution by Era — Secondary Technologies (All Outcomes)",
       subtitle = "Vertical lines = P25, P50, P75",
       x = "Years in queue", y = "Cohort era")
```

---

# 9 · Technology Rankings

```{r rankings, fig.height=6}
# Rank each technology by: came_online median, withdrew median, success rate
rank_df <- df_focus %>%
  group_by(tech) %>%
  summarise(
    n_total       = n(),
    n_co          = sum(outcome == "came_online"),
    n_wd          = sum(outcome == "withdrew"),
    success_rate  = n_co / n_total,
    med_co        = median(tiq_days[outcome == "came_online"], na.rm = TRUE),
    med_wd        = median(tiq_days[outcome == "withdrew"],    na.rm = TRUE),
    p90_co        = quantile(tiq_days[outcome == "came_online"], 0.9, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_co >= 20)

rank_df %>%
  mutate(
    success_pct = paste0(round(100 * success_rate, 1), "%"),
    med_co_yr   = round(med_co / 365.25, 1),
    med_wd_yr   = round(med_wd / 365.25, 1),
    p90_co_yr   = round(p90_co / 365.25, 1)
  ) %>%
  arrange(med_co) %>%
  select(tech, n_total, n_co, success_pct, med_co_yr, med_wd_yr, p90_co_yr) %>%
  kbl(
    caption   = "Technology rankings by queue efficiency",
    col.names = c("Technology","Total N","Came Online","Success Rate",
                  "Median Online (yrs)","Median Withdrew (yrs)","P90 Online (yrs)")
  ) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  column_spec(4, bold = TRUE,
              color = ifelse(
                as.numeric(sub("%","",
                  arrange(rank_df %>% filter(n_co>=20), med_co)$success_rate * 100)) > 30,
                "#2e7d32","#c62828")) %>%
  row_spec(0, bold = TRUE)
```

```{r success-vs-time, fig.height=5.5}
rank_df %>%
  ggplot(aes(x = med_co / 365.25, y = success_rate,
             colour = tech, size = n_total, label = tech)) +
  geom_point(alpha = 0.85) +
  geom_label_repel(size = 3.5, fontface = "bold",
                   box.padding = 0.4, show.legend = FALSE) +
  scale_colour_manual(values = pal_tech, guide = "none") +
  scale_size_continuous(range = c(4, 14),
                        name = "Total filings",
                        labels = comma) +
  scale_x_continuous(labels = function(x) paste0(x,"y")) +
  scale_y_continuous(labels = percent, limits = c(0, 0.6)) +
  geom_hline(yintercept = mean(rank_df$success_rate),
             linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = median(rank_df$med_co / 365.25),
             linetype = "dashed", colour = "grey50") +
  labs(title = "Technology: Success Rate vs. Time to Come Online",
       subtitle = "Upper-left = fast and successful · Lower-right = slow and unlikely to complete\nDashed lines = dataset averages",
       x = "Median years to reach commercial operation",
       y = "Success rate (came online / total filings)",
       caption = "Point size proportional to total number of filings")
```

---

# 10 · Summary by Technology

```{r final-summary}
summary_tbl <- tribble(
  ~Technology, ~`Came Online Trend`, ~`Withdrew Trend`, ~`Key Finding`,
  "Solar",
    "↓ Decreasing (−16%)",
    "↓ Decreasing (−24%)",
    "Both trends declining — improved grid integration + developers exiting faster. Dominates the queue numerically.",
  "Wind",
    "→ Flat (−3%)",
    "↓ Decreasing (−70%)",
    "Completion time stable; withdrawal time collapsed. Wind developers are giving up far earlier than in the 2000s.",
  "Gas",
    "↑ Increasing (+32%) **",
    "↓ Decreasing (−10%)",
    "Only technology with a statistically significant and economically large increase in time to come online.",
  "Battery",
    "Insufficient data (post-2008)",
    "→ Flat (+7%)",
    "Too new for came-online trend. Withdrawal times flat — developers hold on rather than exit early.",
  "Coal",
    "↓ Decreasing (−25%)",
    "→ Flat (+3%)",
    "Fewer projects overall; recent approvals came online faster. Regulatory pressure may be expediting the few that proceed.",
  "Nuclear",
    "↑ Increasing (+17%)",
    "↓ Decreasing (−64%)",
    "Small sample (n=46). Longest median came-online time of any technology at 4.2 years.",
  "Hydro",
    "↓ Decreasing (−66%)",
    "↓ Decreasing (−44%)",
    "Large apparent improvement but very small sample — driven by a few projects.",
  "Offshore Wind",
    "Insufficient data (n=1)",
    "↑ Increasing (+57%)",
    "Withdrawal times growing — projects are persisting longer before abandonment, reflecting high development costs."
)

summary_tbl %>%
  kbl(caption = "Technology-level findings summary") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = TRUE) %>%
  column_spec(2, color = ifelse(grepl("↑", summary_tbl$`Came Online Trend`),
                                "#1565C0", ifelse(grepl("↓", summary_tbl$`Came Online Trend`),
                                "#B71C1C", "grey40")), bold = TRUE) %>%
  column_spec(3, color = ifelse(grepl("↑", summary_tbl$`Withdrew Trend`),
                                "#1565C0", ifelse(grepl("↓", summary_tbl$`Withdrew Trend`),
                                "#B71C1C", "grey40")), bold = TRUE) %>%
  row_spec(0, bold = TRUE) %>%
  row_spec(3, background = "#fff3e0")  # highlight Gas as standout
```

---

*Analysis: 9,256 queue filings, cohorts 2000–2019. Methods: Mann-Kendall trend test, Wilcoxon rank-sum, Kaplan-Meier survival analysis, quantile regression. See `queue_time_analysis.Rmd` for the full dataset-level analysis.*
