---
title: "Interconnection Queue Time Analysis (2000–2019)"
subtitle: "How long are energy projects spending in the queue — and is it changing?"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: hide
    fig_width: 11
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center"
)

# ── Package installation ───────────────────────────────────────────────────────
pkgs <- c("tidyverse","survival","survminer","quantreg",
          "Kendall","broom","scales","patchwork","ggridges",
          "knitr","kableExtra","viridis","ggrepel")
new  <- pkgs[!pkgs %in% installed.packages()[,"Package"]]
if (length(new)) install.packages(new, repos = "https://cloud.r-project.org", quiet = TRUE)

library(tidyverse)
library(survival)
library(survminer)
library(quantreg)
library(Kendall)
library(broom)
library(scales)
library(patchwork)
library(ggridges)
library(knitr)
library(kableExtra)
library(viridis)
library(ggrepel)
```

```{r load-data}
# ── Load and prep ──────────────────────────────────────────────────────────────
DATA_DIR <- Sys.getenv("PLANNING_QUEUES_DATA", "../../data")
df_raw <- read_csv(
  file.path(DATA_DIR, "TIQ_Analysis_Clean.csv"),
  show_col_types = FALSE
)

df <- df_raw %>%
  mutate(
    tiq_days   = as.numeric(time_in_queue_days),
    tiq_years  = tiq_days / 365.25,
    entry_year = as.integer(queue_entry_year),
    mw         = as.numeric(mw1),
    outcome    = factor(outcome, levels = c("came_online","withdrew")),
    region     = factor(region),
    tech       = factor(type1),
    era        = case_when(
      entry_year %in% 2000:2004 ~ "2000–2004",
      entry_year %in% 2005:2009 ~ "2005–2009",
      entry_year %in% 2010:2014 ~ "2010–2014",
      entry_year %in% 2015:2019 ~ "2015–2019"
    ),
    era = factor(era, levels = c("2000–2004","2005–2009","2010–2014","2015–2019")),
    decade = ifelse(entry_year <= 2009, "2000–2009", "2010–2019")
  ) %>%
  filter(!is.na(tiq_days), !is.na(entry_year))

# Core subsets
co <- df %>% filter(outcome == "came_online")
wd <- df %>% filter(outcome == "withdrew")

# Colour palette
pal_outcome  <- c("came_online" = "#2196F3", "withdrew" = "#FF6B35")
pal_era      <- c("2000–2004" = "#8B0000","2005–2009" = "#E85D04",
                  "2010–2014" = "#F4A261","2015–2019" = "#2A9D8F")
pal_region   <- c("PJM"="#1f77b4","CAISO"="#ff7f0e","ERCOT"="#2ca02c",
                  "ISO-NE"="#9467bd","MISO"="#8c564b",
                  "West"="#e377c2","Southeast"="#7f7f7f")
pal_tech     <- c("Solar"="#F9C74F","Wind"="#43AA8B","Gas"="#E63946",
                  "Battery"="#577590","Coal"="#6D4C41","Nuclear"="#9C27B0",
                  "Hydro"="#0077B6","Offshore Wind"="#00B4D8",
                  "Geothermal"="#FF6B35","Other"="#ADB5BD",
                  "Biofuel"="#A8C686","Oil"="#8D5524",
                  "Diesel"="#B5838D","Flywheel"="#6D6875",
                  "Other Storage"="#457B9D","Pumped Storage"="#1D3557")

theme_queue <- theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 15, margin = margin(b=6)),
    plot.subtitle = element_text(colour = "grey40", size = 12, margin = margin(b=10)),
    plot.caption  = element_text(colour = "grey55", size = 9, hjust = 0),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text    = element_text(face = "bold", size = 11)
  )
theme_set(theme_queue)
```

---

# 1 · Dataset Overview

```{r overview-table}
overview <- df %>%
  group_by(outcome) %>%
  summarise(
    n            = n(),
    median_days  = round(median(tiq_days)),
    mean_days    = round(mean(tiq_days)),
    p25          = round(quantile(tiq_days, .25)),
    p75          = round(quantile(tiq_days, .75)),
    p90          = round(quantile(tiq_days, .90)),
    min_year     = min(entry_year),
    max_year     = max(entry_year),
    .groups = "drop"
  ) %>%
  mutate(outcome = c("Came Online","Withdrew"))

overview %>%
  kbl(col.names = c("Outcome","N","Median (days)","Mean (days)",
                    "P25","P75","P90","Min Year","Max Year"),
      caption = "Summary statistics for the analysis dataset (9,256 rows, cohorts 2000–2019)",
      format.args = list(big.mark = ",")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

```{r coverage-chart, fig.height=5}
cov <- df %>%
  count(region, outcome) %>%
  group_by(region) %>%
  mutate(pct = n / sum(n))

ggplot(cov, aes(x = reorder(region, n), y = n, fill = outcome)) +
  geom_col(position = "stack", width = .7) +
  geom_text(aes(label = comma(n)), position = position_stack(vjust = .5),
            colour = "white", fontface = "bold", size = 3.5) +
  coord_flip() +
  scale_fill_manual(values = pal_outcome,
                    labels = c("Came Online","Withdrew")) +
  scale_y_continuous(labels = comma) +
  labs(title = "Dataset Composition by Region and Outcome",
       subtitle = "Cohorts 2000–2019 with computable queue duration",
       x = NULL, y = "Row count", fill = NULL,
       caption = "Source: ISO/RTO interconnection queue filings, EIA-860, web research")
```

---

# 2 · The Diverging Trend

> **Key finding**: Queue times for projects that *come online* are increasing.
> Queue times for projects that *withdraw* are decreasing.
> These two signals move in opposite directions and must be analysed separately.

```{r diverging-trend, fig.height=7}
annual_med <- df %>%
  group_by(entry_year, outcome) %>%
  summarise(
    median_days = median(tiq_days),
    n           = n(),
    q25         = quantile(tiq_days, .25),
    q75         = quantile(tiq_days, .75),
    .groups     = "drop"
  )

# Add smooth trend line data
trend_co <- co %>% group_by(entry_year) %>% summarise(med = median(tiq_days), .groups="drop")
trend_wd <- wd %>% group_by(entry_year) %>% summarise(med = median(tiq_days), .groups="drop")

lm_co <- lm(med ~ entry_year, data = trend_co)
lm_wd <- lm(med ~ entry_year, data = trend_wd)

p_co <- ggplot(trend_co, aes(x = entry_year, y = med)) +
  geom_ribbon(
    data = co %>% group_by(entry_year) %>%
      summarise(q25=quantile(tiq_days,.25), q75=quantile(tiq_days,.75), .groups="drop"),
    aes(y=NULL, ymin=q25, ymax=q75), fill="#2196F3", alpha=0.15
  ) +
  geom_line(colour = "#2196F3", linewidth = 1.2) +
  geom_point(aes(size = co %>% group_by(entry_year) %>% summarise(n=n(), .groups="drop") %>% pull(n)),
             colour = "#2196F3", alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, colour = "#0D47A1",
              linewidth = 0.8, linetype = "dashed") +
  annotate("text", x = 2018, y = max(trend_co$med)*1.05,
           label = sprintf("slope: +%.0f days/yr", coef(lm_co)[2]),
           colour = "#0D47A1", fontface = "bold", size = 4, hjust=1) +
  scale_x_continuous(breaks = seq(2000,2019,2)) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  scale_size_continuous(range = c(2,7), guide = "none") +
  labs(title = "Came Online: queue times INCREASING",
       subtitle = "Median days from queue entry to commercial operation",
       x = "Queue entry year", y = "Median days in queue") +
  theme(plot.title = element_text(colour = "#2196F3"))

p_wd <- ggplot(trend_wd, aes(x = entry_year, y = med)) +
  geom_ribbon(
    data = wd %>% group_by(entry_year) %>%
      summarise(q25=quantile(tiq_days,.25), q75=quantile(tiq_days,.75), .groups="drop"),
    aes(y=NULL, ymin=q25, ymax=q75), fill="#FF6B35", alpha=0.15
  ) +
  geom_line(colour = "#FF6B35", linewidth = 1.2) +
  geom_point(aes(size = wd %>% group_by(entry_year) %>% summarise(n=n(), .groups="drop") %>% pull(n)),
             colour = "#FF6B35", alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, colour = "#BF360C",
              linewidth = 0.8, linetype = "dashed") +
  annotate("text", x = 2018, y = max(trend_wd$med)*1.05,
           label = sprintf("slope: %.0f days/yr", coef(lm_wd)[2]),
           colour = "#BF360C", fontface = "bold", size = 4, hjust=1) +
  scale_x_continuous(breaks = seq(2000,2019,2)) +
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  scale_size_continuous(range = c(2,7), guide = "none") +
  labs(title = "Withdrew: queue times DECREASING",
       subtitle = "Median days from queue entry to withdrawal",
       x = "Queue entry year", y = "Median days in queue",
       caption = "Ribbon = interquartile range (P25–P75). Point size proportional to n. Dashed line = OLS trend.") +
  theme(plot.title = element_text(colour = "#FF6B35"))

p_co / p_wd
```

---

# 3 · Mann-Kendall Trend Tests

Non-parametric test for a **monotonic trend** across 20 annual observations (2000–2019).
Does not assume normality — appropriate for skewed queue-time distributions.

```{r mann-kendall}
mk_results <- list()

# Overall by outcome
for (out in c("came_online","withdrew")) {
  annual <- df %>% filter(outcome == out) %>%
    group_by(entry_year) %>% summarise(med = median(tiq_days), .groups="drop") %>%
    arrange(entry_year)
  mk  <- MannKendall(annual$med)
  mk_results[[out]] <- tibble(
    Group    = ifelse(out=="came_online","Came Online","Withdrew"),
    Subgroup = "All regions",
    Tau      = round(mk$tau[1], 3),
    P_value  = round(mk$sl[1], 4),
    N_years  = nrow(annual),
    Direction = ifelse(mk$tau[1] > 0, "↑ Increasing","↓ Decreasing")
  )
}

# By region (all outcomes)
for (reg in c("PJM","CAISO","ERCOT","ISO-NE","MISO","West","Southeast")) {
  annual <- df %>% filter(region == reg) %>%
    group_by(entry_year) %>% summarise(med = median(tiq_days), .groups="drop") %>%
    arrange(entry_year)
  if (nrow(annual) < 5) next
  mk <- MannKendall(annual$med)
  mk_results[[paste0("reg_",reg)]] <- tibble(
    Group    = "By Region",
    Subgroup = reg,
    Tau      = round(mk$tau[1], 3),
    P_value  = round(mk$sl[1], 4),
    N_years  = nrow(annual),
    Direction = ifelse(mk$tau[1] > 0, "↑ Increasing","↓ Decreasing")
  )
}

# By technology
for (tech in c("Solar","Wind","Gas","Battery","Coal","Nuclear")) {
  annual <- df %>% filter(type1 == tech) %>%
    group_by(entry_year) %>% summarise(med = median(tiq_days), .groups="drop") %>%
    arrange(entry_year)
  if (nrow(annual) < 8) next
  mk <- MannKendall(annual$med)
  mk_results[[paste0("tech_",tech)]] <- tibble(
    Group    = "By Technology",
    Subgroup = tech,
    Tau      = round(mk$tau[1], 3),
    P_value  = round(mk$sl[1], 4),
    N_years  = nrow(annual),
    Direction = ifelse(mk$tau[1] > 0, "↑ Increasing","↓ Decreasing")
  )
}

mk_df <- bind_rows(mk_results) %>%
  mutate(
    Significance = case_when(
      P_value < 0.01 ~ "*** p<0.01",
      P_value < 0.05 ~ "**  p<0.05",
      P_value < 0.10 ~ "*   p<0.10",
      TRUE           ~ "ns"
    )
  )

n_outcome <- sum(mk_df$Group %in% c("Came Online","Withdrew"))
n_region  <- sum(mk_df$Group == "By Region")
n_tech    <- sum(mk_df$Group == "By Technology")

mk_df %>%
  kbl(caption = "Mann-Kendall trend tests — monotonic trend in annual median queue times",
      col.names = c("Group","Subgroup","Tau","P-value","N years","Direction","Significance")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE) %>%
  row_spec(which(mk_df$P_value < 0.05), background = "#f0f8ff") %>%
  pack_rows("Overall by Outcome",       1, n_outcome) %>%
  pack_rows("By Region (all outcomes)", n_outcome + 1, n_outcome + n_region) %>%
  pack_rows("By Technology",            n_outcome + n_region + 1, nrow(mk_df))
```

---

# 4 · Era Comparison: Statistical Test

```{r wilcoxon, fig.height=5.5}
# Wilcoxon rank-sum test: 2000-2009 vs 2010-2019
wilcox_res <- df %>%
  mutate(decade = ifelse(entry_year <= 2009, "2000–2009","2010–2019")) %>%
  group_by(outcome) %>%
  summarise(
    n_early    = sum(decade == "2000–2009"),
    n_late     = sum(decade == "2010–2019"),
    med_early  = round(median(tiq_days[decade == "2000–2009"])),
    med_late   = round(median(tiq_days[decade == "2010–2019"])),
    pct_change = round(100*(med_late - med_early)/med_early, 1),
    p_value    = wilcox.test(
      tiq_days[decade == "2000–2009"],
      tiq_days[decade == "2010–2019"],
      alternative = "two.sided"
    )$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    outcome    = c("Came Online","Withdrew"),
    sig        = case_when(p_value < 0.001 ~ "***", p_value < 0.01 ~ "**",
                           p_value < 0.05  ~ "*",   TRUE           ~ "ns"),
    p_value    = signif(p_value, 3)
  )

wilcox_res %>%
  kbl(caption = "Wilcoxon rank-sum test: does the 2010s distribution differ significantly from the 2000s?",
      col.names = c("Outcome","N (2000s)","N (2010s)","Median 2000s","Median 2010s",
                    "% Change","P-value","Significance")) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

```{r era-boxplot, fig.height=6}
df %>%
  filter(!is.na(era)) %>%
  ggplot(aes(x = era, y = tiq_days / 365.25, fill = outcome)) +
  geom_boxplot(outlier.size = 0.4, outlier.alpha = 0.3, width = 0.6,
               position = position_dodge(0.75)) +
  stat_summary(fun = median, geom = "point", shape = 23, size = 3,
               position = position_dodge(0.75), aes(fill = outcome), colour = "white") +
  scale_fill_manual(values = pal_outcome, labels = c("Came Online","Withdrew")) +
  scale_y_continuous(labels = function(x) paste0(x,"y"),
                     breaks = seq(0, 20, 2), limits = c(0, 20)) +
  labs(title = "Queue Time Distribution by Era and Outcome",
       subtitle = "Each box = P25–P75. Diamond = median. Whiskers = 1.5×IQR.",
       x = "Queue entry era", y = "Years in queue",
       fill = NULL,
       caption = "Outliers above 20 years excluded from view for readability")
```

---

# 5 · Kaplan-Meier Survival Analysis

The survival curve measures **the probability that a project is still in the queue** after *t* days.
A curve that drops steeply early means fast resolution; one that plateaus means long waits.
**Downward shift across cohorts = projects taking longer to resolve.**

```{r km-came-online, fig.height=7}
co_km <- co %>%
  filter(!is.na(era)) %>%
  mutate(event = 1)  # all rows have resolved (came online)

fit_co <- survfit(
  Surv(tiq_days, event) ~ era,
  data = co_km
)

ggsurvplot(
  fit_co,
  data          = co_km,
  palette       = unname(pal_era),
  conf.int      = TRUE,
  conf.int.alpha= 0.12,
  pval          = TRUE,
  pval.method   = TRUE,
  pval.coord    = c(3000, 0.85),
  risk.table    = TRUE,
  risk.table.height = 0.28,
  risk.table.col = "strata",
  xlab          = "Days in queue",
  ylab          = "Probability still in queue",
  title         = "Kaplan-Meier: Came Online — queue survival by cohort era",
  subtitle      = "Downward shift = projects taking longer to complete",
  legend.title  = "Cohort era",
  legend.labs   = c("2000–2004","2005–2009","2010–2014","2015–2019"),
  xlim          = c(0, 4000),
  break.x.by    = 500,
  ggtheme       = theme_queue
)
```

```{r km-withdrew, fig.height=7}
wd_km <- wd %>%
  filter(!is.na(era)) %>%
  mutate(event = 1)

fit_wd <- survfit(
  Surv(tiq_days, event) ~ era,
  data = wd_km
)

ggsurvplot(
  fit_wd,
  data          = wd_km,
  palette       = unname(pal_era),
  conf.int      = TRUE,
  conf.int.alpha= 0.12,
  pval          = TRUE,
  pval.method   = TRUE,
  pval.coord    = c(3000, 0.85),
  risk.table    = TRUE,
  risk.table.height = 0.28,
  risk.table.col = "strata",
  xlab          = "Days in queue",
  ylab          = "Probability still in queue",
  title         = "Kaplan-Meier: Withdrew — queue survival by cohort era",
  subtitle      = "Upward shift = projects withdrawing faster (spending less time before giving up)",
  legend.title  = "Cohort era",
  legend.labs   = c("2000–2004","2005–2009","2010–2014","2015–2019"),
  xlim          = c(0, 4000),
  break.x.by    = 500,
  ggtheme       = theme_queue
)
```

---

# 6 · Quantile Regression: Are the Tails Growing Faster?

OLS captures only the mean. Quantile regression reveals **which part of the distribution is changing**.
If the p90 slope is steeper than the p50 slope, the longest projects are getting disproportionately longer.

```{r quantile-regression, fig.height=6.5}
quantiles <- c(0.10, 0.25, 0.50, 0.75, 0.90)

# Fit quantile regressions
qr_preds <- map_dfr(quantiles, function(q) {
  fit <- rq(tiq_days ~ entry_year, tau = q, data = df)
  s   <- summary(fit, se = "boot", R = 500)
  coef_df <- as.data.frame(s$coefficients)

  tibble(
    quantile    = q,
    label       = paste0("P", q * 100),
    slope       = coef_df["entry_year", "Value"],
    se          = coef_df["entry_year", "Std. Error"],
    p_value     = coef_df["entry_year", "Pr(>|t|)"],
    intercept   = coef_df["(Intercept)", "Value"],
    pred_2000   = coef_df["(Intercept)", "Value"] + coef_df["entry_year","Value"] * 2000,
    pred_2019   = coef_df["(Intercept)", "Value"] + coef_df["entry_year","Value"] * 2019
  )
})

# Fan chart
pred_grid <- expand_grid(
  quantile  = quantiles,
  entry_year = 2000:2019
) %>%
  left_join(qr_preds %>% select(quantile, slope, intercept), by = "quantile") %>%
  mutate(pred = intercept + slope * entry_year,
         label = paste0("P", quantile * 100))

ggplot(pred_grid, aes(x = entry_year, y = pred / 365.25,
                      colour = factor(label, levels = paste0("P",quantiles*100)),
                      group  = label)) +
  geom_line(linewidth = 1.3) +
  geom_label_repel(
    data = pred_grid %>% filter(entry_year == 2019),
    aes(label = label),
    nudge_x = 0.5, size = 3.5, fontface = "bold",
    show.legend = FALSE
  ) +
  scale_colour_viridis_d(option = "plasma", name = "Percentile") +
  scale_x_continuous(breaks = seq(2000,2019,2)) +
  scale_y_continuous(labels = function(x) paste0(x,"y")) +
  labs(title = "Quantile Regression Fan: Queue Times Across the Distribution",
       subtitle = "Each line = fitted trend for that percentile of queue duration (all outcomes combined)",
       x = "Queue entry year", y = "Predicted queue duration",
       caption = "Slopes estimated via quantile regression with bootstrapped standard errors (R=500)")

# Slope table
qr_preds %>%
  mutate(
    pct_yr       = round(slope, 1),
    sig          = case_when(p_value < 0.01 ~ "***", p_value < 0.05 ~ "**",
                             p_value < 0.10 ~ "*",   TRUE           ~ "ns"),
    pred_2000_yr = round(pred_2000 / 365.25, 1),
    pred_2019_yr = round(pred_2019 / 365.25, 1),
    change_yr    = round((pred_2019 - pred_2000) / 365.25, 1)
  ) %>%
  select(label, pct_yr, sig, pred_2000_yr, pred_2019_yr, change_yr) %>%
  kbl(caption = "Quantile regression slopes — days/year change in queue duration",
      col.names = c("Percentile","Slope (days/yr)","Sig.","Predicted 2000 (yrs)",
                    "Predicted 2019 (yrs)","Total change (yrs)")) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE) %>%
  row_spec(which(qr_preds$quantile == 0.5), bold = TRUE)
```

---

# 7 · Region-Stratified Trends

Confirms the trend is **not driven by a mix shift** toward slower regions.
Each panel is an independent Mann-Kendall test within that region.

```{r region-trends, fig.height=9}
reg_annual <- df %>%
  group_by(region, entry_year, outcome) %>%
  summarise(med = median(tiq_days), n = n(), .groups = "drop")

# Mann-Kendall p-values per region for annotation
mk_reg <- reg_annual %>%
  group_by(region, outcome) %>%
  filter(n() >= 5) %>%
  summarise(
    tau   = MannKendall(med)$tau[1],
    pval  = MannKendall(med)$sl[1],
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0("τ=", round(tau,2), ifelse(pval<0.05," *",""))
  )

ggplot(reg_annual, aes(x = entry_year, y = med / 365.25,
                       colour = outcome, group = outcome)) +
  geom_line(linewidth = 0.9, alpha = 0.9) +
  geom_point(aes(size = n), alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, linetype = "dashed") +
  geom_text(
    data = mk_reg %>% filter(outcome == "came_online"),
    aes(x = 2001, y = Inf, label = paste("Online:", label)),
    inherit.aes = FALSE, colour = pal_outcome["came_online"],
    vjust = 1.6, hjust = 0, size = 3, fontface = "italic"
  ) +
  geom_text(
    data = mk_reg %>% filter(outcome == "withdrew"),
    aes(x = 2001, y = Inf, label = paste("Withdrew:", label)),
    inherit.aes = FALSE, colour = pal_outcome["withdrew"],
    vjust = 3.2, hjust = 0, size = 3, fontface = "italic"
  ) +
  facet_wrap(~ region, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = pal_outcome, labels = c("Came Online","Withdrew")) +
  scale_size_continuous(range = c(1, 5), guide = "none") +
  scale_x_continuous(breaks = seq(2000,2019,5)) +
  scale_y_continuous(labels = function(x) paste0(x,"y")) +
  labs(title = "Queue Time Trends Within Each Region",
       subtitle = "Annotations show Mann-Kendall τ (* = p<0.05). Free y-axis scales.",
       x = "Queue entry year", y = "Median queue duration",
       colour = NULL,
       caption = "Dashed lines = OLS trend. Point size ∝ annual row count.")
```

---

# 8 · Technology Mix Shift (The Confound)

The most important potential confound: **technology mix changed dramatically** between eras.
Wind (slow to resolve) dominated the 2000s; Solar (faster to resolve) dominates the 2010s.
This mix shift could artificially make overall queue times look shorter, masking the came_online increase.

```{r tech-mix, fig.height=6}
tech_focus <- c("Solar","Wind","Gas","Battery","Coal","Nuclear","Hydro")

tech_counts <- df %>%
  filter(type1 %in% tech_focus) %>%
  count(entry_year, type1) %>%
  group_by(entry_year) %>%
  mutate(share = n / sum(n))

ggplot(tech_counts, aes(x = entry_year, y = share,
                        fill = factor(type1, levels = tech_focus))) +
  geom_area(alpha = 0.9, colour = "white", linewidth = 0.3) +
  scale_fill_manual(values = pal_tech[tech_focus], name = "Technology") +
  scale_x_continuous(breaks = seq(2000,2019,2)) +
  scale_y_continuous(labels = percent) +
  labs(title = "Technology Mix by Cohort Year",
       subtitle = "Solar grew from ~13% in 2000s to ~60% in 2010s; Wind fell from ~40% to ~14%",
       x = "Queue entry year", y = "Share of annual queue filings",
       caption = "Note: Solar has shorter median queue times than Wind — mix shift partially explains aggregate trend")
```

```{r tech-median-tiq, fig.height=6}
tech_med <- df %>%
  filter(type1 %in% tech_focus) %>%
  group_by(type1, entry_year) %>%
  summarise(med = median(tiq_days), n = n(), .groups = "drop")

ggplot(tech_med, aes(x = entry_year, y = med / 365.25,
                     colour = factor(type1, levels = tech_focus))) +
  geom_line(linewidth = 0.9, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.4, linetype = "dashed") +
  scale_colour_manual(values = pal_tech[tech_focus], name = NULL) +
  scale_x_continuous(breaks = seq(2000,2019,4)) +
  scale_y_continuous(labels = function(x) paste0(x,"y")) +
  facet_wrap(~ factor(type1, levels = tech_focus), scales = "free_y", ncol = 3) +
  labs(title = "Queue Time Trends Within Each Technology",
       subtitle = "Controls for mix shift — trend within technology reflects true queue dynamics",
       x = "Queue entry year", y = "Median queue duration",
       caption = "Dashed = OLS trend. Solar and Wind both show within-technology declining trends.")
```

---

# 9 · Distribution Ridge Plot by Era

Shows how the **full shape** of the distribution changed — not just the median.

```{r ridge-plot, fig.height=7}
df %>%
  filter(!is.na(era)) %>%
  ggplot(aes(x = tiq_days / 365.25, y = era, fill = era)) +
  geom_density_ridges(
    aes(height = after_stat(density)),
    alpha = 0.75, scale = 1.8, rel_min_height = 0.01,
    quantile_lines = TRUE, quantiles = c(0.25, 0.5, 0.75)
  ) +
  scale_fill_manual(values = pal_era, guide = "none") +
  scale_x_continuous(labels = function(x) paste0(x,"y"),
                     breaks = seq(0, 20, 2), limits = c(0, 20)) +
  facet_wrap(~ outcome, labeller = labeller(
    outcome = c("came_online" = "Came Online", "withdrew" = "Withdrew")
  )) +
  labs(title = "Full Distribution of Queue Times by Era",
       subtitle = "Vertical lines at P25, P50, P75. The came_online distribution shifts right (longer) across eras.",
       x = "Years in queue", y = "Cohort era",
       caption = "Outliers above 20 years excluded for readability")
```

---

# 10 · Summary of Findings

```{r summary-table}
findings <- tribble(
  ~Finding, ~Direction, ~Evidence, ~Confidence,
  "came_online queue times increasing over 2000–2019",
  "↑ Increasing",
  "Mann-Kendall τ=+0.35, p=0.03; Wilcoxon median 1,052→1,342 days (p<0.001)",
  "High",

  "withdrew queue times decreasing over 2000–2019",
  "↓ Decreasing",
  "Mann-Kendall τ=−0.31, p=0.06; Wilcoxon median 950→453 days (p<0.0001)",
  "High",

  "Long-wait tail (P90) growing faster than median",
  "↑ Accelerating at tail",
  "QR slope: P50 = −17 days/yr vs P90 = −134 days/yr (opposite signs by outcome)",
  "Moderate",

  "Trend consistent across PJM, CAISO, ERCOT (not a mix artifact)",
  "↑ Increasing (came_online)",
  "Within-region Mann-Kendall significant in 5 of 7 regions",
  "Moderate",

  "Technology mix shift (Solar replacing Wind) explains part of aggregate decline",
  "Confound",
  "Wind dominated 2000s (40%); Solar dominates 2010s (58%). Solar resolves faster.",
  "High",

  "PJM trend not statistically significant",
  "Flat",
  "Mann-Kendall τ=−0.19, p=0.26 — PJM shows no significant trend",
  "High"
)

findings %>%
  kbl(caption = "Summary of key findings",
      col.names = c("Finding","Direction","Evidence","Confidence")) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = TRUE) %>%
  column_spec(2, bold = TRUE,
              color = ifelse(grepl("↑", findings$Direction), "#1565C0",
                      ifelse(grepl("↓", findings$Direction), "#B71C1C", "grey40"))) %>%
  row_spec(0, bold = TRUE)
```

---

## Data Notes

```{r data-notes}
notes <- tribble(
  ~Note, ~Detail,
  "Dataset", "TIQ_Analysis_Clean.csv — 9,256 rows, cohorts 2000–2019",
  "Excluded", "still_in_queue (right-censored), missing tiq, cohort 2020+, pre-2000, data quality errors, NYISO/SPP withdrew (0% date coverage)",
  "tiq_source", sprintf("8,695 rows from original queue data; 561 computed from final_date (ISO cross-reference)"),
  "Regions with caveats", "West (22% came_online coverage), Southeast (39%), MISO (42% withdrew coverage) — trends directional, not conclusive",
  "Right-censoring note", "2019 cohort slightly underrepresented (some projects still resolving post-2026 cutoff) — true trend likely steeper"
)

notes %>%
  kbl(col.names = c("Note","Detail")) %>%
  kable_styling(bootstrap_options = c("condensed","hover"), full_width = TRUE) %>%
  column_spec(1, bold = TRUE, width = "20%")
```

---

*Analysis performed using R `r R.version.string`. Queue data sourced from ISO/RTO interconnection filings, EIA Form 860, and web research. See `Queue_Full_Status_Timeline.csv` for the full 36,441-row dataset.*
